---
title: "MLB Payroll Distribution and Team Performance Analysis (2015-2024)"
author: "Selah Dean"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required packages
library(tidyverse)
library(broom)
library(plm)
library(ineq)     
library(patchwork)
library(kableExtra)
library(corrplot)
library(scales)
library(lme4)
library(patchwork)
library(modelsummary)
library(fixest)

```

## Introduction

This analysis explores the relationship between Major League Baseball (MLB) team payroll distribution and team success over a ten-season period (2015–2024). We examine whether financial constraints affect team performance and whether small-market teams can achieve similar results as big-market teams despite payroll disparities.

## Data Import and Preparation

```{r}

roster_payroll_data <- read.csv(paste0(getwd(), "/datasets/cleaned/roster_payroll_data.csv"))
payroll_with_standings <- read.csv(paste0(getwd(), "/datasets/cleaned/payroll_with_standings.csv"))

```

```{r}
status_summary <- roster_payroll_data %>%
  filter(!is.na(status_simple) &
           !is.na(payroll_salary_adjusted)) %>%
  group_by(team_id, year, status_simple) %>%
  summarize(
    num_players = n(),
    min_salary = min(payroll_salary_adjusted),
    max_salary = max(payroll_salary_adjusted),
    avg_salary = mean(payroll_salary_adjusted),
    total_payroll = sum(payroll_salary_adjusted),
    payroll_gini = ineq(payroll_salary_adjusted, type = "Gini"),
    .groups = "drop"
  ) %>% rbind(
    roster_payroll_data %>%
      filter(!is.na(payroll_salary_adjusted)) %>%
      group_by(team_id, year) %>%
      summarize(
        status_simple = "overall",
        num_players = n(),
        min_salary = min(payroll_salary_adjusted),
        max_salary = max(payroll_salary_adjusted),
        avg_salary = mean(payroll_salary_adjusted),
        total_payroll = sum(payroll_salary_adjusted),
        payroll_gini = ineq(payroll_salary_adjusted, type = "Gini"),
        .groups = "drop"
      )
  ) %>%
  pivot_wider(
    names_from = status_simple,
    values_from = c(num_players, min_salary, max_salary, avg_salary, total_payroll, payroll_gini),
    names_glue = "{status_simple}_{.value}"
  ) %>% rename_with(~ if_else(.x == "team_id", .x,  
                        str_to_lower(str_replace_all(.x, "-", "_")) %>%
                          str_replace("^(.{3})[^_]*_", "\\1_") %>% 
                          str_replace("^pre_([a-z]{3})[^_]*_", "pre_\\1_") %>%
                          str_remove("^ove_"))) %>%
  left_join(
    payroll_with_standings %>% select(team_id, year, league, division, market_size, W, W.L., win_variability),
    by = join_by(team_id, year)
  ) %>%
  rename(wins = W, 
         win_pct = W.L.) %>%
  mutate(
    market_size = case_when(
       market_size >= 5e6 ~ "large",
       market_size < 5e6 & market_size >= 2.5e6 ~ "mid",
       market_size < 2.5e6 ~ "small"
     ),
     market_size = factor(market_size, levels = c("small", "mid", "large")), 
    pre_arb_payroll_pct = pre_arb_total_payroll / total_payroll,
    arb_payroll_pct = arb_total_payroll / total_payroll, 
    vet_payroll_pct = vet_total_payroll / total_payroll, 
    pre_arb_player_pct = pre_arb_num_players / num_players, 
    arb_player_pct = arb_num_players / num_players, 
    vet_player_pct = vet_num_players / num_players
  )

status_summary_wo2020 <- status_summary %>% filter(year != 2020)

```

## Exploratory Data Analysis

### Descriptive Statistics

```{r descriptive_stats}
# Year-by-year summary
yearly_stats <- status_summary %>%
  group_by(year) %>%
  summarize(
    mean_wins = mean(wins),
    sd_wins = sd(wins),
    min_wins = min(wins),
    max_wins = max(wins),
    
    mean_payroll = mean(total_payroll),
    sd_payroll = sd(total_payroll),
    min_payroll = min(total_payroll),
    max_payroll = max(total_payroll),
    mean_gini = mean(payroll_gini), 
    
    mean_players = mean(num_players),
    sd_players = sd(num_players),
    min_players = min(num_players),
    max_players = max(num_players),
    
    mean_pre_arb_players = mean(pre_arb_num_players),
    sd_pre_arb_players = sd(pre_arb_num_players),
    min_pre_arb_players = min(pre_arb_num_players),
    max_pre_arb_players = max(pre_arb_num_players),
    
    mean_arb_players = mean(arb_num_players),
    sd_arb_players = sd(arb_num_players),
    min_arb_players = min(arb_num_players),
    max_arb_players = max(arb_num_players),
    
    mean_vet_players = mean(vet_num_players),
    sd_vet_players = sd(vet_num_players),
    min_vet_players = min(vet_num_players),
    max_vet_players = max(vet_num_players),
    
    mean_pre_arb_pct = mean(pre_arb_payroll_pct),
    sd_pre_arb_pct = sd(pre_arb_payroll_pct),
    min_pre_arb_pct = min(pre_arb_payroll_pct),
    max_pre_arb_pct = max(pre_arb_payroll_pct),
    
    mean_arb_pct = mean(arb_payroll_pct),
    sd_arb_pct = sd(arb_payroll_pct),
    min_arb_pct = min(arb_payroll_pct),
    max_arb_pct = max(arb_payroll_pct),
    
    mean_vet_pct = mean(vet_payroll_pct),
    sd_vet_pct = sd(vet_payroll_pct),
    min_vet_pct = min(vet_payroll_pct),
    max_vet_pct = max(vet_payroll_pct),
  ) %>%
  mutate(
    across(ends_with("_payroll"), ~ round(. / 1e6, 2))
  )

kable(yearly_stats, 
      caption = "League-Wide Trends by Year",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

team_stats <- status_summary %>%
  filter(year != 2020) %>%
  group_by(team_id) %>%
  summarize(
    mean_wins = mean(wins),
    sd_wins = sd(wins),
    min_wins = min(wins),
    max_wins = max(wins),
    
    mean_payroll = mean(total_payroll),
    sd_payroll = sd(total_payroll),
    min_payroll = min(total_payroll),
    max_payroll = max(total_payroll),
    mean_gini = mean(payroll_gini), 
    
    mean_players = mean(num_players),
    sd_players = sd(num_players),
    min_players = min(num_players),
    max_players = max(num_players),
    
    mean_pre_arb_players = mean(pre_arb_num_players),
    sd_pre_arb_players = sd(pre_arb_num_players),
    min_pre_arb_players = min(pre_arb_num_players),
    max_pre_arb_players = max(pre_arb_num_players),
    
    mean_arb_players = mean(arb_num_players),
    sd_arb_players = sd(arb_num_players),
    min_arb_players = min(arb_num_players),
    max_arb_players = max(arb_num_players),
    
    mean_vet_players = mean(vet_num_players),
    sd_vet_players = sd(vet_num_players),
    min_vet_players = min(vet_num_players),
    max_vet_players = max(vet_num_players),
    
    mean_pre_arb_pct = mean(pre_arb_payroll_pct),
    sd_pre_arb_pct = sd(pre_arb_payroll_pct),
    min_pre_arb_pct = min(pre_arb_payroll_pct),
    max_pre_arb_pct = max(pre_arb_payroll_pct),
    
    mean_arb_pct = mean(arb_payroll_pct),
    sd_arb_pct = sd(arb_payroll_pct),
    min_arb_pct = min(arb_payroll_pct),
    max_arb_pct = max(arb_payroll_pct),
    
    mean_vet_pct = mean(vet_payroll_pct),
    sd_vet_pct = sd(vet_payroll_pct),
    min_vet_pct = min(vet_payroll_pct),
    max_vet_pct = max(vet_payroll_pct),
  ) %>%
  mutate(
    across(ends_with("_payroll"), ~ round(. / 1e6, 2))
  )

kable(team_stats, 
      caption = "Trends by Team",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

### Correlation Matrix

```{r correlation_matrix}
# Calculate correlation matrix for key variables
cor_matrix <- status_summary %>%
  filter(year != 2020) %>% 
  select(-team_id, -year, -league, -division, -market_size) %>%
  cor()


corrplot(cor_matrix, type = 'upper',  tl.col = "black",
         tl.cex = 0.6)
```

### Visualizations

#### Overall Payroll 
```{r}
team_payroll_by_year <- ggplot(status_summary %>% filter(year !=2020), 
                               aes(x=total_payroll, y=wins)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "red", size=0.5) + 
  facet_wrap(vars(year), ncol=3) + 
  xlab("Team Total Payroll") + ylab("Regular Season Wins") + 
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_minimal() + ggtitle("Team Payroll vs Wins (2015-2024)") + 
  theme(plot.title = element_text(hjust = 0.5))

team_payroll_by_year

ggsave("figures/team_payroll_by_year.png", team_payroll_by_year, 
       width = 8, height = 8, dpi = 300, bg = "white")
```

#### Payroll Distribution by Player Category

```{r}
# First, get the team order from the payroll distribution data
team_order_data <- status_summary %>%
  group_by(team_id, market_size) %>%
  summarize(
    pre_arb = mean(pre_arb_payroll_pct),
    .groups = 'drop'
  ) %>%
  group_by(market_size) %>%
  mutate(team_order = reorder(team_id, pre_arb)) %>%
  select(market_size, team_id, team_order)

# First plot - Payroll distribution
plot1 <- status_summary %>%
  group_by(team_id, market_size) %>%
  summarize(
    pre_arb = mean(pre_arb_payroll_pct),
    arb = mean(arb_payroll_pct),
    vet = mean(vet_payroll_pct),
    .groups = 'drop'
  ) %>%
  # Join with the team order data
  left_join(team_order_data, by = c("team_id", "market_size")) %>%
  pivot_longer(cols = c(pre_arb, arb, vet), 
               names_to = "category", 
               values_to = "percentage") %>%
  mutate(
    category = factor(category, 
                     levels = c("pre_arb", "arb", "vet"),
                     labels = c("Pre-Arbitration", "Arbitration", "Veteran"))
  ) %>%
  ggplot(aes(x = team_order, y = percentage, fill = category)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE)) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Payroll Distribution",
    subtitle = "(2015-2024 Average)",
    x = "Team",
    y = "Percentage of Payroll",
    fill = "Player Category"
  ) +
  facet_grid(rows = vars(market_size), scales = "free_y", space = "free_y") +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

# Second plot - Player distribution
plot2 <- status_summary %>%
  group_by(team_id, market_size) %>%
  summarize(
    pre_arb = mean(pre_arb_num_players / num_players),
    arb = mean(arb_num_players / num_players),
    vet = mean(vet_num_players / num_players),
    .groups = 'drop'
  ) %>%
  # Join with the team order data to use the same ordering
  left_join(team_order_data, by = c("team_id", "market_size")) %>%
  pivot_longer(cols = c(pre_arb, arb, vet), 
               names_to = "category", 
               values_to = "percentage") %>%
  mutate(
    category = factor(category, 
                     levels = c("pre_arb", "arb", "vet"),
                     labels = c("Pre-Arbitration", "Arbitration", "Veteran"))
  ) %>%
  ggplot(aes(x = team_order, y = percentage, fill = category)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE)) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Player Distribution",
    subtitle = "(2015-2024 Average)",
    x = "",
    y = "Percentage of Players",
    fill = "Player Category"
  ) +
  facet_grid(rows = vars(market_size), scales = "free_y", space = "free_y") +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

# Combine the plots side by side with shared legend
payroll_player_dist <- plot1 + plot2 + 
  plot_layout(guides = "collect") & # Share the legend between plots
  theme(legend.position = "bottom")  # Position the legend at the bottom

payroll_player_dist

# Display the combined plot
ggsave("figures/payroll_player_dist.png", payroll_player_dist, 
       width = 12, height = 8, dpi = 300, bg = "white")
```

#### Consistency Analysis

```{r consistency_analysis}
# Calculate consistency metrics for each team
consistency <- status_summary %>%
  filter(year != 2020) %>%
  group_by(team_id, market_size) %>%
  summarize(
    avg_wins = mean(wins),
    sd_wins = sd(wins),
    avg_payroll_gini = mean(payroll_gini), 
    sd_payroll_gini = sd(payroll_gini), 
    avg_payroll = mean(total_payroll),
    sd_payroll = sd(total_payroll),
    min_payroll = min(total_payroll), 
    max_payroll = max(total_payroll), 
    avg_pre_arb = mean(pre_arb_payroll_pct),
    avg_arb = mean(arb_payroll_pct), 
    avg_vet = mean(vet_payroll_pct),
    .groups = 'drop'
  )

# Plot average wins vs. standard deviation of wins
avg_wins_consistency <- ggplot(consistency, aes(x = avg_wins, y = sd_wins, color = market_size)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size=0.5) +
  ggrepel::geom_text_repel(aes(label = team_id), size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(
    #title = "Average Wins vs. Win Consistency",
    x = "Average Wins per Season",
    y = "Standard Deviation of Wins",
    color = "Market Size"
  ) + 
  scale_y_continuous(
    limits = c(0, max(consistency$sd_wins + 1, na.rm = TRUE)),
    breaks = seq(0, max(consistency$sd_wins + 1, na.rm = TRUE), by = 5)
  ) +
  theme(plot.title = element_text(hjust = 0.5))

# Plot average payroll vs. standard deviation of wins
avg_payroll_consistency <- ggplot(consistency, aes(x = avg_payroll, y = sd_wins, color = market_size)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size=0.5) +
  ggrepel::geom_text_repel(aes(label = team_id), size = 3, max.overlaps = 10) +
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_minimal() +
  labs(
    #title = "Average Total Payroll vs. Win Consistency",
    x = "Average Total Payroll per Season",
    y = "Standard Deviation of Wins",
    color = "Market Size"
  ) + 
    scale_y_continuous(
    limits = c(0, max(consistency$sd_wins + 1, na.rm = TRUE)),
    breaks = seq(0, max(consistency$sd_wins + 1, na.rm = TRUE), by = 5)
  ) +
  theme(plot.title = element_text(hjust = 0.5))

# Plot payroll stability vs. performance stability
sd_payroll_consistency <- ggplot(consistency, aes(x = sd_payroll, y = sd_wins, color = market_size)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(aes(label = team_id), size = 3, max.overlaps = 10) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size=0.5) +
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_minimal() +
  labs(
    #title = "Payroll Consistency vs. Win Consistency",
    x = "Standard Deviation of Total Payroll",
    y = "Standard Deviation of Wins",
    color = "Market Size"
  ) + 
    scale_y_continuous(
    limits = c(0, max(consistency$sd_wins + 1, na.rm = TRUE)),
    breaks = seq(0, max(consistency$sd_wins + 1, na.rm = TRUE), by = 5)
  ) +
  theme(plot.title = element_text(hjust = 0.5))



# Plot average wins vs. standard deviation of wins
avg_pre_arb_pct_consistency <- ggplot(consistency, aes(x = avg_pre_arb, y = sd_wins, color = market_size)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size=0.5) +
  ggrepel::geom_text_repel(aes(label = team_id), size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(
    title = "Average Pre-Arb Payroll% vs. Win Consistency",
    x = "Average Pre-Arb Payroll% per Season",
    y = "Standard Deviation of Wins",
    color = "Market Size"
  ) + 
    scale_y_continuous(
    limits = c(0, max(consistency$sd_wins + 1, na.rm = TRUE)),
    breaks = seq(0, max(consistency$sd_wins + 1, na.rm = TRUE), by = 5)
  ) +
  theme(plot.title = element_text(hjust = 0.5))

# Plot average wins vs. standard deviation of wins
avg_arb_pct_consistency <- ggplot(consistency, aes(x = avg_arb, y = sd_wins, color = market_size)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size=0.5) +
  ggrepel::geom_text_repel(aes(label = team_id), size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(
    title = "Average Arb Payroll% vs. Win Consistency",
    x = "Average Arb Payroll% per Season",
    y = "Standard Deviation of Wins",
    color = "Market Size"
  ) + 
    scale_y_continuous(
    limits = c(0, max(consistency$sd_wins + 1, na.rm = TRUE)),
    breaks = seq(0, max(consistency$sd_wins + 1, na.rm = TRUE), by = 5)
  ) +
  theme(plot.title = element_text(hjust = 0.5))

# Plot average wins vs. standard deviation of wins
avg_vet_pct_consistency <- ggplot(consistency, aes(x = avg_vet, y = sd_wins, color = market_size)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size=0.5) +
  ggrepel::geom_text_repel(aes(label = team_id), size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(
    title = "Average Vet Payroll% vs. Win Consistency",
    x = "Average Vet Payroll% per Season",
    y = "Standard Deviation of Wins",
    color = "Market Size"
  ) + 
    scale_y_continuous(
    limits = c(0, max(consistency$sd_wins + 1, na.rm = TRUE)),
    breaks = seq(0, max(consistency$sd_wins + 1, na.rm = TRUE), by = 5)
  ) +
  theme(plot.title = element_text(hjust = 0.5))

# Plot average wins vs. standard deviation of wins
avg_payroll_gini_consistency <- ggplot(consistency, aes(x = avg_payroll_gini, y = sd_wins, color = market_size)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size=0.5) +
  ggrepel::geom_text_repel(aes(label = team_id), size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(
    title = "Average Payroll Gini Index vs. Win Consistency",
    x = "Average Payroll Gini Index per Season",
    y = "Standard Deviation of Wins",
    color = "Market Size"
  ) + 
    scale_y_continuous(
    limits = c(0, max(consistency$sd_wins + 1, na.rm = TRUE)),
    breaks = seq(0, max(consistency$sd_wins + 1, na.rm = TRUE), by = 5)
  ) +
  theme(plot.title = element_text(hjust = 0.5))

# Plot average wins vs. standard deviation of wins
sd_payroll_gini_consistency <- ggplot(consistency, aes(x = sd_payroll_gini, y = sd_wins, color = market_size)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size=0.5) +
  ggrepel::geom_text_repel(aes(label = team_id), size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(
    title = "Average Payroll Gini Index vs. Win Consistency",
    x = "Average Payroll Gini Index per Season",
    y = "Standard Deviation of Wins",
    color = "Market Size"
  ) + 
    scale_y_continuous(
    limits = c(0, max(consistency$sd_wins + 1, na.rm = TRUE)),
    breaks = seq(0, max(consistency$sd_wins + 1, na.rm = TRUE), by = 5)
  ) +
  theme(plot.title = element_text(hjust = 0.5))


avg_wins_consistency <- avg_wins_consistency +
  theme(axis.title.y = element_blank())

avg_payroll_consistency <- avg_payroll_consistency +
  theme(axis.title.y = element_blank())

sd_payroll_consistency <- sd_payroll_consistency +
  theme(axis.title.y = element_blank())

# Create shared y-axis label using text annotation
y_label <- ggplot() +
  theme_void() +
  annotate("text", x = 1, y = 0.5, angle = 90, label = "Standard Deviation of Wins", size = 4)

wins_payroll_consistency <- (y_label | avg_wins_consistency | avg_payroll_consistency | sd_payroll_consistency) +
  plot_layout(widths = c(0.05, 1, 1, 1), guides = "collect") +
  plot_annotation(
    title = "Wins and Payroll vs Win Consistency (2015-2024)",
    theme = theme(plot.title = element_text(size = 18,  hjust = 0.5))
  ) &
  theme(legend.position = "bottom")

wins_payroll_consistency

ggsave("figures/wins_payroll_consistency.png", wins_payroll_consistency, width = 12, height = 4, dpi = 300, bg = "white")
```


```{r}
# Calculate correlations
avg_wins_corr <- cor(consistency$avg_wins, consistency$sd_wins, use = "complete.obs")
avg_payroll_corr <- cor(consistency$avg_payroll, consistency$sd_wins, use = "complete.obs")
sd_payroll_corr <- cor(consistency$sd_payroll, consistency$sd_wins, use = "complete.obs")

# Format the correlations for display
avg_wins_corr_text <- paste("r =", round(avg_wins_corr, 2))
avg_payroll_corr_text <- paste("r =", round(avg_payroll_corr, 2))
sd_payroll_corr_text <- paste("r =", round(sd_payroll_corr, 2))

# Modify each plot to include correlation
avg_wins_consistency <- avg_wins_consistency +
  annotate("text", x = max(consistency$avg_wins, na.rm = TRUE) - 0.05 * diff(range(consistency$avg_wins, na.rm = TRUE)), 
           y = max(consistency$sd_wins, na.rm = TRUE) - 0.1 * diff(range(consistency$sd_wins, na.rm = TRUE)), 
           label = avg_wins_corr_text, hjust = 1, size = 4.5)

avg_payroll_consistency <- avg_payroll_consistency +
  annotate("text", x = max(consistency$avg_payroll, na.rm = TRUE) - 0.05 * diff(range(consistency$avg_payroll, na.rm = TRUE)), 
           y = max(consistency$sd_wins, na.rm = TRUE) - 0.1 * diff(range(consistency$sd_wins, na.rm = TRUE)), 
           label = avg_payroll_corr_text, hjust = 1, size = 4.5)

sd_payroll_consistency <- sd_payroll_consistency +
  annotate("text", x = max(consistency$sd_payroll, na.rm = TRUE) - 0.05 * diff(range(consistency$sd_payroll, na.rm = TRUE)), 
           y = max(consistency$sd_wins, na.rm = TRUE) - 0.1 * diff(range(consistency$sd_wins, na.rm = TRUE)), 
           label = sd_payroll_corr_text, hjust = 1, size = 4.5)

# Then use the modified plots in your final combined plot
wins_payroll_consistency <- (y_label | avg_wins_consistency | avg_payroll_consistency | sd_payroll_consistency) +
  plot_layout(widths = c(0.05, 1, 1, 1), guides = "collect") +
  plot_annotation(
    title = "Wins and Payroll vs Win Consistency (2015-2024)",
    theme = theme(plot.title = element_text(size = 18,  hjust = 0.5))
  ) &
  theme(legend.position = "bottom")

wins_payroll_consistency

ggsave("figures/wins_payroll_consistency_with_r.png", wins_payroll_consistency, width = 12, height = 4, dpi = 300, bg = "white")
```
#### Market Size Comparisons

```{r}
pvals <- status_summary %>%
  filter(year != 2020) %>%
  select(market_size, wins, total_payroll, payroll_gini, pre_arb_payroll_pct, 
         arb_payroll_pct, vet_payroll_pct) %>%
  pivot_longer(-market_size, names_to = "metric", values_to = "value") %>%
  group_by(metric) %>%
  do(tidy(aov(value ~ market_size, data = .))) %>%
  filter(term == "market_size") %>%
  select(metric, p.value) %>%
  mutate(
    label = case_when(
      p.value < 0.001 ~ "< 0.001",
      TRUE ~ sprintf("%.3f", p.value)
    ),
    metric = factor(metric,
                   levels = c("wins", "total_payroll", "payroll_gini", 
                              "pre_arb_payroll_pct", "arb_payroll_pct", "vet_payroll_pct"),
                   labels = c("Wins", "Payroll", "Payroll Gini Index", 
                              "Pre-Arb Payroll%", "Arb Payroll%", "Vet Payroll%")), 
    facet_label = paste0(metric, "\n(p = ", label, ")")
  ) 


plot_data <- status_summary %>%
  filter(year != 2020) %>%
  select(market_size, wins, total_payroll, payroll_gini, 
         pre_arb_payroll_pct, arb_payroll_pct, vet_payroll_pct) %>%
  pivot_longer(-market_size, names_to = "metric", values_to = "value") %>%
  mutate(
    metric = factor(metric,
                   levels = c("wins", "total_payroll", "payroll_gini", 
                              "pre_arb_payroll_pct", "arb_payroll_pct", "vet_payroll_pct"),
                   labels = c("Wins", "Payroll", "Payroll Gini Index", 
                              "Pre-Arb Payroll%", "Arb Payroll%", "Vet Payroll%")), 
  ) %>%
  left_join(pvals, by = "metric") %>%
  arrange(metric) %>%
  mutate(facet_label = factor(facet_label, levels = unique(facet_label)))

metrics_by_market_size <- ggplot(plot_data, aes(x = market_size, 
                                                y = value, fill = market_size)) +
  geom_boxplot() +
  facet_wrap(~facet_label, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Metrics by Market Size (2015–2024)",
    x = "Market Size",
    y = "Value",
    fill = "Market Size"
  ) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

metrics_by_market_size

ggsave("figures/metrics_by_market_size.png", metrics_by_market_size, 
       width = 8, height = 6, dpi = 300, bg = "white")

```

## Regressions 
```{r}
status_summary_wo2020 <- status_summary_wo2020 %>%
  mutate(
    total_payroll_m = total_payroll / 1e6
  )
```

### Simple Linear Model (Total Payroll vs Wins)

```{r}


model_base      <- feols(wins ~ total_payroll_m, data = status_summary_wo2020)
model_team_fe   <- feols(wins ~ total_payroll_m | team_id, data = status_summary_wo2020)
model_year_fe   <- feols(wins ~ total_payroll_m | year, data = status_summary_wo2020)
model_both_fe   <- feols(wins ~ total_payroll_m | team_id + year, data = status_summary_wo2020)

model_log_base      <- feols(wins ~ log(total_payroll_m), data = status_summary_wo2020)
model_log_team_fe   <- feols(wins ~ log(total_payroll_m) | team_id, data = status_summary_wo2020)
model_log_year_fe   <- feols(wins ~ log(total_payroll_m) | year, data = status_summary_wo2020)
model_log_both_fe   <- feols(wins ~ log(total_payroll_m) | team_id + year, data = status_summary_wo2020)

# Named list of models
models <- list(
  "Baseline"              = model_base,
  "Team FE"               = model_team_fe,
  "Year FE"               = model_year_fe,
  "Team + Year FE"        = model_both_fe,
  "Baseline"        = model_log_base,
  "Team FE"         = model_log_team_fe,
  "Year FE"         = model_log_year_fe,
  "Team + Year FE"  = model_log_both_fe
)

# Use msummary to allow grouping and clean formatting
simple_model_summary <- modelsummary(
  models,
  statistic = "std.error",
  stars = TRUE, 
  #gof_omit = "r.squared.within|Lik|BIC",
  gof_map = c("nobs", "r.squared", "adj.r.squared", "aic", "rmse"), 
  output = "kableExtra", 
  title = "Effect of Total Payroll on Regular Season Wins"
)

simple_model_summary %>% kable_classic(full_width = F, html_font = "Cambria") %>%
   add_header_above(c(" " = 1, "Linear" = 4, "Log" = 4))





```


#### Visualizing Team Fixed Effects 

```{r}
team_effects <- fixef(model_log_both_fe)$team_id
team_effects_df <- data.frame(
  team_id = names(team_effects),
  effect = as.numeric(team_effects)
)

# Center the effects (subtract the mean) to show relative performance
team_effects_df$effect_centered <- team_effects_df$effect - mean(team_effects_df$effect)

team_effects_df <- team_effects_df %>% left_join(status_summary_wo2020 %>% select(team_id, market_size) %>% distinct(), by="team_id")


# Now plot the centered effects
team_fixed_effects <- ggplot(team_effects_df, aes(x = reorder(team_id, effect_centered), 
                           y = effect_centered,
                           fill = market_size)) +  # Use market_size for fill color
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  coord_flip() +
  labs(
    title = "Team Fixed Effects",
    x = "Team",
    y = "Effect on Wins (Relative to Average Team)",
    fill = "Market Size"  # Legend title
  ) +
  theme_minimal() + 
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))

team_fixed_effects

ggsave("figures/team_fixed_effects.png", team_fixed_effects, width = 6, height = 5, dpi = 300, bg = "white")
```

### Payroll Allocation Model 
```{r}
model_base      <- lm(wins ~ pre_arb_payroll_pct + arb_payroll_pct + vet_payroll_pct -
                        1 ,
                      data = status_summary_wo2020)
model_payroll      <- lm(
  wins ~ pre_arb_payroll_pct + arb_payroll_pct + vet_payroll_pct + total_payroll_m - 1,
  data = status_summary_wo2020
)
model_interactions <- lm(
  wins ~ pre_arb_payroll_pct + arb_payroll_pct + vet_payroll_pct + total_payroll_m +
    arb_payroll_pct:total_payroll_m + vet_payroll_pct:total_payroll_m - 1,
  data = status_summary_wo2020
)

model_market_size <- lm(
  wins ~ pre_arb_payroll_pct + arb_payroll_pct + vet_payroll_pct + market_size - 1,
  data = status_summary_wo2020
)

model_base_variability      <- lm(win_variability ~ pre_arb_payroll_pct + arb_payroll_pct + vet_payroll_pct -
                        1 ,
                      data = status_summary_wo2020)
model_payroll_variability      <- lm(
  win_variability ~ pre_arb_payroll_pct + arb_payroll_pct + vet_payroll_pct + total_payroll_m - 1,
  data = status_summary_wo2020
)
model_interactions_variability <- lm(
  win_variability ~ pre_arb_payroll_pct + arb_payroll_pct + vet_payroll_pct + total_payroll_m +
    arb_payroll_pct:total_payroll_m + vet_payroll_pct:total_payroll_m - 1,
  data = status_summary_wo2020
)

model_market_size_variability <- lm(
  win_variability ~ pre_arb_payroll_pct + arb_payroll_pct + vet_payroll_pct + market_size - 1,
  data = status_summary_wo2020
)

# Named list of models
models <- list(
  "Baseline"              = model_base,
  "Total Payroll"        = model_payroll,
  "Market Size"          = model_market_size,
  "Payroll Interaction"          = model_interactions, 
  "Baseline"              = model_base_variability,
  "Total Payroll"        = model_payroll_variability,
  "Market Size"          = model_market_size_variability,
  "Payroll Interaction"          = model_interactions_variability
)

modelsummary(
  models,
  statistic = "std.error",
  stars = TRUE, 
  gof_omit = "Lik|BIC", 
  output = "kableExtra",
  escape = FALSE, 
  title = "Effect of Payroll Allocations on Wins and Win Variability"
) %>% kable_classic(full_width = F, html_font = "Cambria") %>%
   add_header_above(c(" " = 1, "Wins" = 4, "Win Variability" = 4))
```
