---
title: "Initial Data Analysis"
author: "Selah Dean"
date: "`r Sys.Date()`"
output: pdf_document 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("C:/Users/sgdea/ECON399")
library(tidyverse)
library(naniar)
library(scales)
library(gridExtra)
library(grid)
library(stringr)
library(stringi)
library(stringdist)
```

# Data Cleaning 

## Import Data Sets 
```{r}
# Spotrac payroll data  
payroll_data <- read.csv("payroll_data_from_2015.csv") %>% select(-X)
payroll_ranks <- read.csv("payroll_ranks_from_2015.csv") %>% select(-X)

# Baseball Reference 40 man roster data
forty_man <- read.csv("bb_ref_forty_man_roster_from_2015.csv") %>% select(-X)

# Baseball Reference standings data
standings <- read.csv("standings_from_2015.csv") %>% select(-X) %>%
  filter(Tm != "Average")

# File that contains each team, there bbref abbr, league, and division info
team_league_map <- read.csv("team_league_map.csv")
```

## Match League Info for Each Dataset 
```{r}
# Join standings data with team_league_map 
standings <- inner_join(standings, team_league_map, by = join_by(Tm == team))

# Join forty man data with team_league_map
# Only include the newest names to match spotrac
forty_man <- inner_join(
  forty_man,
  team_league_map %>%
    filter(team != "Cleveland Indians" &
             team != "Los Angeles Angels of Anaheim"),
  by = join_by(team == team_id)
) %>%
  rename(team_id = team, team = team.y)

# modify the format of the team name to match with spotrac 
team_league_map <- team_league_map %>% mutate(
  team = str_to_lower(team) %>%
    gsub(" ", "-", .) %>%
    gsub("\\.", "", .) %>%
    gsub("-of-anaheim", "", .) %>%
    gsub("indians", "guardians", .) %>%
    gsub("oakland-", "", .)
)  %>% distinct()

# Join payroll rank data on team_league_map
payroll_ranks <- inner_join(payroll_ranks, team_league_map, by = join_by(team))

# Join payroll data on team_league_map
payroll_data <- inner_join(payroll_data, team_league_map, by = join_by(team))

```

## General Data Cleaning
```{r}
# Clean payroll rank so amount and rank are numeric 
payroll_ranks <- payroll_ranks %>% mutate(
    amount = as.numeric(gsub("[\\$,]", "", amount)), 
    rank = as.numeric(gsub("[a-zA-Z]", "", rank))
  )

# Clean payroll data so salaries are numeric 
# Replace any empty strings with NA
# Simplify the roster column to not include year or the word "Payroll"
# This should match the designations found in the payroll_ranks
payroll_data <- payroll_data %>%
  mutate(across(
    c(
      payroll_salary,
      payroll_salary_adjusted,
      base_salary,
      signing_bonus,
      incentives_likely,
      incentives_unlikely
    ),
    ~ as.numeric(gsub("[\\$,]", "", na_if(., "-")))
  )) %>% 
  mutate(across(where(is.character), ~ na_if(.x, ""))) %>%
  mutate(
    roster = str_remove(roster, "^\\d{4} ") %>%  
                          str_remove(" Payroll$")
  )

# Determine total payroll for each team 
total_payroll <- payroll_ranks %>% 
  filter(player_type == "Total Adjusted Payroll Allocations")

# Clean 40 man data
# rename X.1 column to all_star
# Make Yrs numeric 
forty_man <- forty_man %>%
  rename(
    all_star = X.1
  ) %>%
  mutate(
    all_star = case_when(
      all_star == "All-Star" ~ TRUE, 
      TRUE ~ FALSE
    ), 
    Yrs = as.numeric(gsub("[a-zA-Z]", "", Yrs))
  )
  
```


### Handle Duplicate Players in Payroll 
Trades and releases complicate the payroll data slightly causing players to appear in different places on the payroll. 

First wanted to confirm that the payroll data in for each player matches the overall team data. 

```{r}
payroll_summary <- payroll_data %>% 
  group_by(team, year, roster) %>%
  summarize(
    total_payroll_adj = sum(payroll_salary_adjusted, na.rm = T),
    .groups = "drop"
  )

unmatched <- left_join(payroll_summary, payroll_ranks, 
                  by = join_by(roster == player_type, team, year)) %>%
  filter(total_payroll_adj != amount) %>%
  mutate(diff = total_payroll_adj - amount )

unique(unmatched$roster)
```
Determined that the adjusted values match except with Minor and Retained payroll. Spotrac defines Retained payroll as leftover salaries from players who have been traded or released, or buyouts for declined options. 

```{r}

diff_neg <- unmatched %>% filter(diff < 0)
diff_pos <- unmatched %>% filter(diff > 0)

unique(diff_neg$roster)
unique(diff_pos$roster)

```
All of the negative differences (adding the players' adjusted salaries together results in a higher payroll than found in the ranks) come from retained salaries. Similarly all of the positive differences come from minor league salaries. Need to dive more into this issue. 

There is also the issue of duplicates where a player shows up multiple times on the payroll. 

NOTE: When doing matching discovered that in 2024 the Twins had two different players with the name Diego Castillo. One is a relief pitcher and the other is a 2nd baseman. Initially grouped by position by this dropped duplicates that should not have been dropped due to slight differences in position (i.e. RP vs SP or LF vs RF). Instead going to created a simplified version of position (pitcher, hitter, or two-way (special for Ohtani)) to group by. 

```{r}
# Add a column that designates player as pitcher, hitter, or two-way 
payroll_data <- payroll_data %>%
  mutate(
    pos_simple = case_when(
      pos == "SP/DH" ~ "two-way", 
      pos %in% c("RP", "SP", "SP1", "P", "CL") ~ "pitcher", 
      TRUE ~ "hitter"
    )
  )
  
duplicates <- payroll_data %>%
  group_by(player, team, year, pos_simple) %>%
  filter(n() > 1) %>%
  ungroup()

duplicate_summary <- duplicates %>% 
  group_by(player, team, year, pos_simple) %>%
  summarize(
    num = n(), 
    .groups = "drop"
  )

print(nrow(duplicates))
print(nrow(duplicate_summary))
```
There are 384 duplicate players (same name, team, and year) accounting for 778 rows in the payroll dataframe. 


The duplicates appear to come from when the player is on different roster designations throughout the season resulting in split contracts. To get the player's overall salary from the team I am going to add the duplicates together. For the other values, I want to keep it the same if they are the same in the duplicates, keep the non-null, and combine any remaining unmatched values. 

```{r}
# Function to combine duplicates into one row
combine_duplicates <- function(x) {
  if (all(is.na(x))) {
    return(NA)  # return NA if all values are NA
  } else if (is.numeric(x)) {
    return(sum(x, na.rm = TRUE))  # sum any numeric values
  } else if (is.character(x)) {
    return(paste(unique(na.omit(x)), collapse = "/"))  # concat diff values
  } else {
    return(unique(na.omit(x)))  # keep the value if the same between rows 
  }
}

payroll_data <- payroll_data %>% 
  group_by(player, team, year, pos_simple) %>%
  summarize(across(everything(), combine_duplicates), .groups = "drop")
```

394 rows where removed from the payroll data dataframe which is what was expected (778 - 384).

## Determine Position based on Appearances 

Also want to create the simple position column for the forty man data (will help with the issue different players on same team with same name). Also realized that Shohei Ohtani's position wouldn't be right since he pitches in fewer games than he hits. Added a case to handle this. 

```{r}
# Assign a position to each player based on the most games played at a position
# For pitchers determine is starting or relief pitcher 
forty_man <- forty_man %>%
  mutate(
    position = apply(select(., P:DH) %>% select(-OF), 1, function(row)
      names(row)[which.max(row)]), 
    position = case_when(
      position == "P" & GS >= 0.8 * G ~ "SP", 
      position == "P" ~ "RP", 
      TRUE ~ position
    ), 
    position = gsub("X", "", position), 
    # manually fix Ohtani's position, don't love this solution 
    position = case_when(
      Name == "Shohei Ohtani" ~ "SP/DH", 
      TRUE ~ position
    ),
    pos_simple = case_when(
      position == "SP/DH" ~ "two-way", 
      position %in% c("RP", "SP") ~ "pitcher", 
      TRUE ~ "hitter"
    )
  )
```

## Match Players from 40 man Roster to Payroll Data 

```{r}
# Function to match names
match_players <- function(roster, payroll) {
  # Step 1: Clean the names by removing accents
  roster <- roster %>% mutate(
    player_name_clean = stri_trans_general(Name, "Latin-ASCII")
    )
  payroll <- payroll %>% mutate(
    player_name_clean = stri_trans_general(player, "Latin-ASCII")
    )
  
  # Step 2: Initial exact match on the cleaned names
  exact_matches <- roster %>%
    inner_join(payroll, by = 
                 join_by(player_name_clean, team_id, year, pos_simple))
  
  # Step 3: Find unmatched rows in roster and payroll
  unmatched_roster <- roster %>%
    anti_join(exact_matches, by = "player_name_clean")
  unmatched_payroll <- payroll %>%
    anti_join(exact_matches, by = "player_name_clean")
  
  # Step 4: Fuzzy matching for unmatched names in roster
  fuzzy_matches <- unmatched_roster %>%
    rowwise() %>%
    mutate(best_match_index = which.min(
      stringdist(player_name_clean, payroll$player_name_clean, method = "jw")
    ),
    best_dist = min(
      stringdist(player_name_clean, payroll$player_name_clean, method = "jw")
    )) %>%
    mutate(best_match_name = payroll$player_name_clean[best_match_index]) %>%
    left_join(payroll, 
              by = join_by(best_match_name == player_name_clean, 
                           team_id, year, pos_simple))
  
  
  
  # Step 5: Determine the remaining unmatched names in payroll
  matched_names <- c(exact_matches$player, fuzzy_matches$player)
  unmatched_payroll <- unmatched_payroll %>%
    filter(!player %in% matched_names)
  
  # Step 6: Combine exact matches, fuzzy matches, and remaining players from payroll
  # The remaining players are those who did not make a major league appearance for the team
  final_result <- exact_matches %>%
    bind_rows(fuzzy_matches %>% 
                select(-best_match_index, -best_dist, -best_match_name)) %>%
    bind_rows(unmatched_payroll)
  
  return(final_result)
}

roster_payroll_data <- match_players(forty_man, payroll_data)
```


```{r}
not_in_forty_man <- roster_payroll_data %>% filter(is.na(Name))

not_in_forty_man_active <- not_in_forty_man %>% filter(grepl("Active", roster))

unique(not_in_forty_man_active$player)
```
Summary of why these players weren't matched: 

- Name Issue 
  - C.J. Edwards (Carl Edwards Jr.)
  - Donnie Walton (Donovan Walton)
  - Frank German: (Franklin German)
  - Jon Heasley: (Jonathan Heasley)
  - Josh H. Smith (Josh Smith)
  - Leonardo Jimenez (Leo Jimenez)
  - Manuel Pina (Manny Pina)
  - Melvin Upton (B.J. Upton)
  - Michael Brosseau (Mike Brosseau)
  - Michael Shawaryn (Mike Shawaryn)
  - Mike Siani (Michael Siani)
  - Nathan Eaton (Nate Eaton)
  - Raffy Lopez (Rafael Lopez)
  
  
  
- International free agent that never played pro ball (shows up as one year deal on spotrac)
  - Carlos Mendoza (this is not the now manager)
  - Dahian Santos
  - Darwin Rosario 
  - Diosfran Cabeza
  - Francisco Ventura
  - Jose Astudillo
  - Jose Leanez
  - Marino Castillo
  - Miguel Villarroel
  - Roger Contreras
  - Wilder Perez
  - Winston Santos

- Didn't play in the majors 
  - Erik Davis
  - Jake Smith 
  - Josmil Pinto
  - Kyle Skipworth 
  - Lorenzo Quintana
  - Melvin Adon 
  - Ryan Gusto
  - Thad Weber
  - Yorman Rodríguez

- Other
  - James Jones: Spotrac has him listed as pitcher when he was actually a hitter 
  - Javier Guerra: Was a SS in 2018 and then became a relief pitcher in 2019  
  - Russell Brewer: Played in minors from 2010-2013, not sure what's up with 2015 

```{r}
roster <- forty_man %>% mutate(
    player_name_clean = stri_trans_general(Name, "Latin-ASCII")
)
payroll <- payroll_data %>% mutate(
    player_name_clean = stri_trans_general(player, "Latin-ASCII")
)
  
unmatched_players <- anti_join(roster, payroll, by=join_by(player_name_clean))

# Function to extract last name while removing suffixes
extract_last_name <- function(full_name) {
  name_parts <- str_split(full_name, " ")[[1]]
  
  # Remove common suffixes
  suffixes <- c("Jr.", "Sr.", "II", "III", "IV", "V", "HOF")
  name_parts <- name_parts[!name_parts %in% suffixes]
  
  # Assume last non-suffix part is the last name
  last_name <- tail(name_parts, 1)
  
  # If there is a hyphen extract the last part
  if (str_detect(last_name, "-")){
    last_name <- str_split(last_name, "-")[[1]][2]
  }
  return(last_name)
}

# Function to extract the first letter of the first name
extract_first_initial <- function(full_name) {
  name_parts <- str_split(full_name, " ")[[1]]
  first_initial <- substr(name_parts[1], 1, 1)
  return(first_initial)
}

# Precompute last names and first initials
unmatched_players <- unmatched_players %>%
  mutate(
    last_name = sapply(player_name_clean, extract_last_name),
  )

payroll <- payroll %>%
  mutate(
    last_name = sapply(player_name_clean, extract_last_name),
  )

# Join on last name, first initial, team, and year
matched_candidates <- unmatched_players %>%
  inner_join(payroll, by = c("last_name", "team_id", "year"), 
             suffix = c("_unmatched", "_payroll")) %>%
  group_by(player_name_clean_unmatched, team_id, year) %>%
  mutate(
    dist = stringdist(player_name_clean_unmatched, player_name_clean_payroll, method = "jw")
  ) %>%
  slice_min(dist, n = 1, with_ties = FALSE) %>%  # Get the best match per unmatched player
  ungroup() %>%
  select(player_name_clean_unmatched, player_name_clean_payroll, team_id, year, dist)

# Manually fix the 2 players that weren't matched correctly 
matched_candidates <- matched_candidates %>% 
  mutate(
    player_name_clean_payroll = case_when(
      player_name_clean_unmatched == "B.J. Upton" ~ "Melvin Upton", 
      player_name_clean_unmatched == "Luis Jimenez" ~ NA, 
      TRUE ~ player_name_clean_payroll
    )
  )

remaining_unmatched_players <- anti_join(unmatched_players, matched_candidates, 
                       by = join_by(player_name_clean == player_name_clean_unmatched))

# Merge results back to unmatched_players
unmatched_players <- unmatched_players %>%
  left_join(matched_candidates, 
            by = c("player_name_clean" = "player_name_clean_unmatched", 
                   "team_id", "year"))

# Rename columns for clarity
unmatched_players <- unmatched_players %>%
  rename(
    best_match_name = player_name_clean_payroll,
    best_dist = dist
  )


```


462 unmatched instances (coming from 120 players). When forcing last name to match and then computing fuzzy search, gives 446 matches (for 110 players). 

Incorrect matches: 
- B.J. Upton -> Justin Upton 
- Luis Jimenez -> Cesar Jimenez

16 remaining unmatched instances. All of these players correctly don't have matches since for whatever reason Spotrac doesn't have their salaries. 

```{r}
# Function to clean name 
clean_name <- function(Name) {
  # Convert encoding to Latin-ASCII
  Name <- stri_trans_general(Name, "Latin-ASCII")
  
  # Remove suffixes
  suffixes <- c("Jr.", "Sr.", "II", "III", "IV", "V", "HOF")
  suffix_pattern <- paste0("\\s*(", paste(suffixes, collapse = "|"), ")\\s*$")
  Name <- stri_replace_all_regex(Name, suffix_pattern, "")
  #Name <- stri_replace_all_regex(Name, pattern, "")

  
  # Remove periods and dashes
  Name <- stri_replace_all_regex(Name, "[.]", "")
  
  Name <- stri_replace_all_regex(Name, "[-]", " ")
  
  # Convert to title case 
  Name <- str_to_title(Name)
  
  Name <- gsub('[\t\n]','',Name)
  
  return(Name)
}

# Function to extract last name while removing suffixes
extract_last_name <- function(full_name) {
  name_parts <- str_split(full_name, " ")[[1]]
  last_name <- tail(name_parts, 1)
  
  return(last_name)
}

match_players <- function(roster, payroll) {
  # Step 1: Clean the names by removing accents and create column for last name
  roster <- roster %>% mutate(
    player_name_clean = sapply(Name, clean_name), 
    last_name = sapply(player_name_clean, extract_last_name)
    )
  
  payroll <- payroll %>% mutate(
    player_name_clean = sapply(player, clean_name), 
    last_name = sapply(player_name_clean, extract_last_name)
    )
  
  # Step 2: Initial exact match on the cleaned names
  exact_matches <- roster %>%
    inner_join(payroll, by = 
                 join_by(player_name_clean, team_id, year, pos_simple), 
               suffix = c("", ""))
  
  # Step 3: Find unmatched rows in roster and payroll
  unmatched_roster <- roster %>%
    anti_join(exact_matches, by = join_by(player_name_clean, team_id, year, pos_simple))
  
  unmatched_payroll <- payroll %>%
    anti_join(exact_matches, by = join_by(player_name_clean, team_id, year, pos_simple))
  
  # Step 4: Fuzzy matching for unmatched names in roster
  last_name_matches <- unmatched_roster %>%
    inner_join(
      unmatched_payroll,
      by =join_by(last_name, team_id, year, pos_simple),
      suffix = c("", "")
    ) 
  
  combined_matches <- exact_matches %>%
    bind_rows(last_name_matches)
  
  unmatched_roster_2 <- unmatched_roster %>%
    anti_join(combined_matches, by = join_by(Name, team_id, year, pos_simple))
  
  unmatched_payroll_2 <- unmatched_payroll %>%
    anti_join(combined_matches, by = join_by(player, team_id, year, pos_simple))
  
  wo_position_matches <- unmatched_roster_2 %>%
    inner_join(unmatched_payroll_2, by = 
                 join_by(player_name_clean, team_id, year), 
               suffix = c("_bbref", "_spotrac")) %>% 
    rename(team = team_spotrac) %>%
    select(-contains("_spotrac"), -team_bbref) %>%
    rename_with(~ str_replace(.x, "_bbref$", ""), ends_with("_bbref"))
  
  combined_matches_2 <- combined_matches %>%
    bind_rows(wo_position_matches)
  
  unmatched_roster_3 <- unmatched_roster_2 %>%
    anti_join(combined_matches_2, by = join_by(Name, team_id, year, pos_simple))
  
  unmatched_payroll_3 <- unmatched_payroll_2 %>%
    anti_join(combined_matches_2, by = join_by(player, team_id, year, pos_simple))

  # Step 6: Combine exact matches, fuzzy matches, and remaining players from payroll
  # The remaining players are those who did not make a major league appearance for the team
  final_result <- combined_matches_2 %>%
    bind_rows(unmatched_roster_3)
  
  return(final_result)
}

roster_payroll_data <- match_players(forty_man, payroll_data)

```

Since I am matching datasets from two different sources the naming of some of the players is off. After cleaning the name (removing any accents, suffixes, and special characters), 15141 out of 15579 players from the forty man dataset are able to be matched on the payroll dataset. This leaves 438 unmatched players. Next matching on just last name for the remaining unmatched players gives another 255, leaving 183 unmatched players. Then matching without including position gives another 50 matches. This leaves 133 players that still don't have matches. Need to further investigate. It appears that some just don't have the salary information on spotrac.  


# Initial Plots 

## Payroll vs Team Performance 
```{r}
# Combine payroll data and standings
payroll_with_standings <- inner_join(
  total_payroll, standings, 
  by = join_by(team_id, year, league, division)
) 

# Payroll Rank vs Regular Season Rank
ggplot(payroll_with_standings, aes(x=rank, y=Rk)) + 
  geom_point() + 
  xlab("Payroll Rank") + ylab("Regular Season Rank") + 
  xlim(1, 30) + ylim(1,30) + 
  theme_bw() + ggtitle("Payroll vs Team Performance (Regular Season Rank)") + 
  theme(plot.title = element_text(hjust = 0.5))

# Payroll Rank vs Win Percentage
ggplot(payroll_with_standings, aes(x=rank, y=W.L.)) + 
  geom_point() + 
  xlab("Payroll Rank") + ylab("Win Percentage") + 
  xlim(1, 30) + 
  theme_bw() + ggtitle("Payroll vs Team Performance (Win Percentage)") + 
  theme(plot.title = element_text(hjust = 0.5))

# Payroll Amount vs Win Percentage
ggplot(payroll_with_standings, aes(x=amount, y=W.L.)) + 
  geom_point() + 
  xlab("Payroll Rank") + ylab("Win Percentage") + 
  #xlim(1, 30) + 
  theme_bw() + ggtitle("Payroll vs Team Performance (Win Percentage)") + 
  theme(plot.title = element_text(hjust = 0.5))

# Payroll Rank vs Regular Season Rank split by Season 
ggplot(payroll_with_standings, aes(x=rank, y=Rk)) + 
  geom_point() + 
  xlab("Payroll Rank") + ylab("Regular Season Rank") + 
  xlim(1, 30) + ylim(1,30) + 
  theme_bw() + facet_wrap(vars(year), ncol=5) + 
  ggtitle("Payroll vs Team Performance (Regular Season Rank) by Season") + 
  theme(plot.title = element_text(hjust = 0.5))

# Payroll Amount vs Regular Season Rank split by Season 
ggplot(payroll_with_standings, aes(x=amount, y=W.L.)) + 
  geom_point() + 
  xlab("Payroll Rank") + ylab("Regular Season Rank") + 
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M"), guide = guide_axis(angle = 45)) + 
  theme_bw() + facet_wrap(vars(year), ncol=5) + 
  ggtitle("Payroll vs Team Performance (Regular Season Rank) by Season") + 
  theme(plot.title = element_text(hjust = 0.5))

# Payroll Rank vs Win Percentage split by Season 
ggplot(payroll_with_standings, aes(x=rank, y=W.L.)) + 
  geom_point() + 
  xlab("Payroll Rank") + ylab("Win Percentage") + 
  xlim(1, 30) + 
  theme_bw() + facet_wrap(vars(year), ncol=5) + 
  ggtitle("Payroll vs Team Performance (Win Percentage) by Season") + 
  theme(plot.title = element_text(hjust = 0.5))

# Payroll Amount vs Wins split by Season 
ggplot(payroll_with_standings %>% filter(year !=2020), aes(x=amount, y=W)) + 
  geom_point() + 
  xlab("Payroll") + ylab("Win Percentage") + 
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_bw() + facet_wrap(vars(year), ncol=5) + 
  ggtitle("Payroll vs Team Performance (Win Percentage) by Season") + 
  theme(plot.title = element_text(hjust = 0.5))

diff <- payroll_with_standings %>% group_by(year) %>%
  summarize(
    payroll_gap = max(amount) - min(amount), 
    win_gap = max(W) - min(W), 
  )

ggplot(diff, aes(x = year, y = payroll_gap)) + geom_point()

ggplot(diff, aes(x = year, y = win_gap)) + geom_point()


ggarrange(ggplot(diff, aes(x = year, y = payroll_gap)) + geom_point(), ggplot(diff, aes(x = year, y = win_gap)) + geom_point())
```

```{r}
ggplot(payroll_with_standings, aes(x = year, y = amount)) + geom_boxplot(aes(group = year))

```
## Payroll by Year

```{r}
team_payroll_by_year <- total_payroll %>% group_by(year) %>%
  summarize(
    avg_team_payroll = mean(amount), 
    max_team_payroll = max(amount), 
    min_team_payroll = min(amount), 
    .groups = "drop"
  )

team_payroll_by_year_by_league <- total_payroll %>% group_by(league, year) %>%
  summarize(
    avg_team_payroll = mean(amount), 
    max_team_payroll = max(amount), 
    min_team_payroll = min(amount), 
    .groups = "drop"
  )

# Avg, Max, Min League Payroll by Season 
ggplot(team_payroll_by_year, aes(x = factor(year))) + 
  geom_point(aes(y = avg_team_payroll, color = "Average Payroll")) + 
  geom_line(aes(y = avg_team_payroll, color = "Average Payroll", group = 1)) + 
  geom_point(aes(y = max_team_payroll, color = "Max Payroll")) + 
  geom_line(aes(y = max_team_payroll, color = "Max Payroll", group = 1)) + 
  geom_point(aes(y = min_team_payroll, color = "Min Payroll")) + 
  geom_line(aes(y = min_team_payroll, color = "Min Payroll", group = 1)) + 
  scale_color_manual(values = c("Average Payroll" = "blue", 
                                "Max Payroll" = "red", 
                                "Min Payroll" = "green")) +
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_bw() + 
  xlab("Season") + 
  ylab("Team Payroll") + 
  labs(color = "Payroll Type") + 
  ggtitle("League Payroll by Season") + 
  theme(plot.title = element_text(hjust = 0.5)) 

team_payroll_by_year_by_league_long <- team_payroll_by_year_by_league %>%
  pivot_longer(
    cols = c(avg_team_payroll, max_team_payroll, min_team_payroll), # Columns to convert
    names_to = "payroll_type",  # New column for variable names
    values_to = "payroll_value" # New column for values
  )

ggplot(team_payroll_by_year_by_league_long, aes(x = factor(year), group = interaction(league, payroll_type))) +
  geom_point(aes(y = payroll_value, color = payroll_type)) +
  geom_line(aes(y = payroll_value, color = payroll_type, linetype = league)) +
  scale_color_brewer(palette = "Set1") +  # Use color for league distinction
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) + # Different line types for avg, max, min
  scale_y_continuous(labels = scales::label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Season") +
  ylab("Team Payroll") +
  labs(color = "", linetype = "League") +
  ggtitle("League Payroll by Season") +
  theme(plot.title = element_text(hjust = 0.5))


```

## Payroll by Position 

```{r}

payroll_by_pos <- roster_payroll_data %>%
  #filter(grepl("Active Roster", roster) | grepl("Injured List", roster)) %>%
  group_by(year, position) %>%
  summarize(
    max_salary = max(payroll_salary, na.rm = TRUE),
    min_salary = min(payroll_salary, na.rm = TRUE),
    avg_salary = mean(payroll_salary, na.rm = TRUE), 
    median_salary = median(payroll_salary, na.rm = TRUE), 
    num_players = n(), 
    .groups = "drop"
  )

payroll_by_pos_by_league <- roster_payroll_data %>%
  group_by(league, year, position) %>%
  summarize(
    max_salary = max(payroll_salary, na.rm = TRUE),
    min_salary = min(payroll_salary, na.rm = TRUE),
    avg_salary = mean(payroll_salary, na.rm = TRUE), 
    median_salary = median(payroll_salary, na.rm = TRUE), 
    num_players = n(), 
    .groups = "drop"
  )

payroll_by_pos_general <- roster_payroll_data %>%
  mutate(
    pos_general = case_when(
      position %in% c("1B", "2B", "SS", "3B") ~ "INF", 
      position %in% c("LF", "CF", "RF") ~ "OF",
      TRUE ~ position
    )
  ) %>% 
  group_by(year, pos_general) %>%
  summarize(
    max_salary = max(payroll_salary, na.rm = TRUE),
    min_salary = min(payroll_salary, na.rm = TRUE),
    avg_salary = mean(payroll_salary, na.rm = TRUE), 
    median_salary = median(payroll_salary, na.rm = TRUE), 
    num_players = n(), 
    .groups = "drop"
  )

payroll_by_pos_by_league_general <- roster_payroll_data %>%
  mutate(
    pos_general = case_when(
      position %in% c("1B", "2B", "SS", "3B") ~ "INF", 
      position %in% c("LF", "CF", "RF") ~ "OF",
      TRUE ~ position
    )
  ) %>% 
  group_by(year, league, pos_general) %>%
  summarize(
    max_salary = max(payroll_salary, na.rm = TRUE),
    min_salary = min(payroll_salary, na.rm = TRUE),
    avg_salary = mean(payroll_salary, na.rm = TRUE), 
    median_salary = median(payroll_salary, na.rm = TRUE), 
    num_players = n(), 
    .groups = "drop"
  )


ggplot(payroll_by_pos_general, aes(x = pos_general, y = num_players)) + 
  geom_bar(stat="identity") + facet_wrap(vars(year), ncol=5) + theme_bw() + 
  xlab("Position") + 
  ylab("Number of Players") + 
  scale_x_discrete(guide = guide_axis(angle = 45)) + 
  ggtitle("Players by Position") + 
  theme(plot.title = element_text(hjust = 0.5)) 

payroll_by_pos_general_long <- payroll_by_pos_general %>%
  pivot_longer(
    cols = c(avg_salary, max_salary, min_salary), # Columns to convert
    names_to = "salary_type",  # New column for variable names
    values_to = "salary_value" # New column for values
  )

payroll_by_pos_by_league_general_long <- payroll_by_pos_by_league_general %>%
  pivot_longer(
    cols = c(avg_salary, max_salary, min_salary), # Columns to convert
    names_to = "salary_type",  # New column for variable names
    values_to = "salary_value" # New column for values
  )

# ggplot(payroll_by_pos_general_long %>% filter(pos_general != "SP/DH" & salary_type != "min_salary"), aes(x = factor(year), group = interaction(pos_general, salary_type))) +
#   geom_point(aes(y = salary_value, color = pos_general)) +
#   geom_line(aes(y = salary_value, color = pos_general, linetype = salary_type)) +
#   scale_color_brewer(palette = "Dark2") + 
#   scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
#   theme_bw() +
#   xlab("Season") +
#   ylab("Salary") +
#   labs(color = "Position", linetype = "Salary") +
#   ggtitle("Salary by Position by Season") +
#   theme(plot.title = element_text(hjust = 0.5))

ggplot(payroll_by_pos_general %>% filter(pos_general != "SP/DH"), aes(x = factor(year))) + 
  geom_point(aes(y = avg_salary, color = factor(pos_general))) + 
  geom_line(aes(y = avg_salary, color = factor(pos_general), group = factor(pos_general))) +
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_bw() + 
  xlab("Season") + 
  ylab("Salary") + 
  labs(color = "Position") + 
  ggtitle("Average Salary by Season") + 
  theme(plot.title = element_text(hjust = 0.5)) 

ggplot(payroll_by_pos_general %>% filter(pos_general != "SP/DH"), aes(x = factor(year))) + 
  geom_point(aes(y = max_salary, color = factor(pos_general))) + 
  geom_line(aes(y = max_salary, color = factor(pos_general), group = factor(pos_general))) +
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_bw() + 
  xlab("Season") + 
  ylab("Salary") + 
  labs(color = "Position") + 
  ggtitle("Highest Salary by Season") + 
  theme(plot.title = element_text(hjust = 0.5)) 

ggplot(payroll_by_pos_by_league_general_long %>% filter(salary_type == "avg_salary" & pos_general != "SP/DH"), aes(x = factor(year), group = interaction(league, pos_general))) +
  geom_point(aes(y = salary_value, color = pos_general)) +
  geom_line(aes(y = salary_value, color = pos_general, linetype = league)) +
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Season") +
  ylab("Salary") +
  labs(color = "", linetype = "League") +
  ggtitle("League Average Salary by Position by Season") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(payroll_by_pos_by_league_general_long %>% filter(salary_type == "max_salary" & pos_general != "SP/DH"), aes(x = factor(year), group = interaction(league, pos_general))) +
  geom_point(aes(y = salary_value, color = pos_general)) +
  geom_line(aes(y = salary_value, color = pos_general, linetype = league)) +
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Season") +
  ylab("Salary") +
  labs(color = "", linetype = "League") +
  ggtitle("League Highest Salary by Position by Season") +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
test <- roster_payroll_data %>% group_by(position, team_id, year) %>% slice_max(payroll_salary_adjusted)

test_24 <- test %>% filter(year == "2024") %>% 
  select(Name, Age, G, WAR, position, payroll_salary_adjusted, roster, team_id, year)

nats_24 <- roster_payroll_data %>% filter(year == 2024 & team_id == "WSN") %>% select(Name, Age, G, WAR, position, payroll_salary_adjusted, roster, team_id, year) %>% mutate(payroll_salary_adjusted = dollar(payroll_salary_adjusted))


test_24_nats <- test_24 %>% filter(team_id == "WSN")
```


```{r}
multiple_team_players <- roster_payroll_data %>%
  group_by(Name, year, pos_simple) %>%
  filter(n() > 1) %>%
  ungroup()

roster_payroll_data_overall <-  roster_payroll_data %>% 
  mutate(Yrs = as.character(Yrs), Age = as.character(Age)) %>% 
  group_by(Name, DoB, year, pos_simple) %>%
  summarize(across(everything(), combine_duplicates), .groups = "drop")

age_payroll_summary <- roster_payroll_data_overall %>%
  group_by(Age) %>%
  summarize(
    max_salary = max(payroll_salary, na.rm = TRUE),
    min_salary = min(payroll_salary, na.rm = TRUE),
    avg_salary = mean(payroll_salary, na.rm = TRUE), 
    num_players = n(), 
    .groups = "drop"
  )

roster_payroll_data_overall_clean <- roster_payroll_data_overall %>% 
  filter(payroll_salary_adjusted >= 0) #%>%
  #mutate(Age = as.numeric(Age))

ggplot(roster_payroll_data_overall_clean, aes(x=Age, y=payroll_salary_adjusted)) + 
  #geom_boxplot(outlier.shape = NA) +  # Boxplots without outliers
  #geom_jitter(width = 0.2, alpha = 0.5, size = 1) + 
  geom_point() + 
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_bw()

```

```{r}
roster_payroll_data_summary <- roster_payroll_data %>%
  group_by(team_id, year, pos_simple) %>%
  summarize(
    payroll = sum(payroll_salary_adjusted, na.rm = TRUE), 
    .groups = "drop"
  )


roster_payroll_data_summary <- roster_payroll_data %>%
  group_by(team_id, year, pos_simple) %>%
  summarize(
    payroll = sum(payroll_salary_adjusted, na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  group_by(team_id, year) %>%
  mutate(payroll_proportion = payroll / sum(payroll)) %>%
  ungroup() 


roster_payroll_data_summary %>% 
  filter(pos_simple == "hitter") %>% 
  slice_max(payroll_proportion, n=10)

roster_payroll_data_summary %>% 
  filter(pos_simple == "hitter") %>% 
  slice_min(payroll_proportion, n=10)

roster_payroll_data_summary %>% 
  filter(pos_simple == "pitcher") %>% 
  slice_max(payroll_proportion, n=10)

roster_payroll_data_summary %>% 
  filter(pos_simple == "pitcher") %>% 
  slice_min(payroll_proportion, n=10)
```

```{r}
#library(gridExtra)
#library(ggpubr) 


# Convert tables to data frames for gridExtra
top_hitters_df <- roster_payroll_data_summary %>% 
  filter(pos_simple == "hitter") %>% 
  slice_max(payroll_proportion, n=10) %>%
  select(-pos_simple, -payroll) %>%
  mutate(
    payroll_proportion = paste0(round(payroll_proportion * 100, 2),"%")
  ) %>% 
  left_join(standings %>% select(team_id, year, W), by = join_by(team_id, year)) %>% 
  rename(Team = team_id, Season = year, `% Payroll` = payroll_proportion, Wins = W)

bottom_hitters_df <- roster_payroll_data_summary %>% 
  filter(pos_simple == "hitter") %>% 
  slice_min(payroll_proportion, n=10) %>%
  select(-pos_simple, -payroll) %>%
  mutate(
    payroll_proportion = paste0(round(payroll_proportion * 100, 2),"%")
  )  %>% 
  left_join(standings %>% select(team_id, year, W), by = join_by(team_id, year)) %>% 
  rename(Team = team_id, Season = year, `% Payroll` = payroll_proportion, Wins = W)

top_pitchers_df <- roster_payroll_data_summary %>% 
  filter(pos_simple == "pitcher") %>% 
  slice_max(payroll_proportion, n=10) %>%
  select(-pos_simple, -payroll) %>%
  mutate(
    payroll_proportion = paste0(round(payroll_proportion * 100, 2),"%")
  )  %>% 
  left_join(standings %>% select(team_id, year, W), by = join_by(team_id, year)) %>% 
  rename(Team = team_id, Season = year, `% Payroll` = payroll_proportion, Wins = W)

bottom_pitchers_df <- roster_payroll_data_summary %>% 
  filter(pos_simple == "pitcher") %>% 
  slice_min(payroll_proportion, n=10) %>%
  select(-pos_simple, -payroll) %>%
  mutate(
    payroll_proportion = paste0(round(payroll_proportion * 100, 2),"%")
  )  %>% 
  left_join(standings %>% select(team_id, year, W), by = join_by(team_id, year)) %>% 
  rename(Team = team_id, Season = year, `% Payroll` = payroll_proportion, Wins = W)


# Function to create a table with theme_bw and smaller font
create_table_plot <- function(df, title) {
    ggtexttable(df, rows = NULL, theme = ttheme("classic", tbody.style = tbody_style(size = 10))) %>% 
    tab_add_title(text = title, face = "bold", padding = unit(0.5, "line"))
}

# Create table plots with titles
top_hitters_plot <- create_table_plot(top_hitters_df, "Top Hitter Payroll %")
bottom_hitters_plot <- create_table_plot(bottom_hitters_df, "Bottom Hitter Payroll %")
top_pitchers_plot <- create_table_plot(top_pitchers_df, "Top Pitcher Payroll %")
bottom_pitchers_plot <- create_table_plot(bottom_pitchers_df, "Bottom Pitcher Payroll %")

# Arrange in a 2x2 grid
grid.arrange(top_hitters_plot, bottom_hitters_plot,
             top_pitchers_plot, bottom_pitchers_plot,
             ncol = 2)

top_hitters_plot

```

```{r}
ggplot(roster_payroll_data, aes(x=G)) + geom_density()

```



```{r}

test <- roster_payroll_data %>% 
  mutate(
    salary_m = payroll_salary_adjusted / 1000000, 
    war_dollar = WAR / salary_m, 
  ) %>% 
  #filter(G >= 30) %>% 
  #slice_min(war_dollar, n=20) %>%
  select(Name, Age, Yrs, team_id, year, G, position, WAR, payroll_salary_adjusted, salary_m, war_dollar) 


ggplot(test, aes(x=payroll_salary_adjusted, y=WAR)) + geom_point()

%>% group_by(team_id, year) %>%
  summarize(
    avg_war_dollar = mean(war_dollar, na.rm = T), 
    min_war_dollar = min(war_dollar, na.rm = T), 
    max_war_dollar = max(war_dollar, na.rm = T), 
    .groups="drop"
  ) %>% 
  left_join(standings %>% select(team_id, year, W), by = join_by(team_id, year))


```