---
title: "Modeling"
author: "Selah Dean"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("C:/Users/sgdea/ECON399")
library(tidyverse)
library(ggpubr)
library(naniar)
library(scales)
library(gridExtra)
library(grid)
library(stringr)
library(stringi)
library(stringdist)
library(deaR)
```

# Notes on Papers

## [Is Winning Everything? A Data Envelopment Analysis of Major League Baseball and the National Football League](https://journals.sagepub.com/doi/epdf/10.1177/1527002503254047)

-   Compares the efficiency of MLB (NL and AL) vs NFL
-   Player salaries as input and on-field-performance as output
    -   Useful results even is a franchise owner’s entire economic
        outcome is not captured
-   A franchise can be inefficient for a few different reasons
    -   Poor management skills

        -   Inaccurately predicts future performance and overspends

    -   Bad luck

        -   Injury prone players

    -   League may reward a large-market franchise with more revenue

        -   Can spend more on player salaries to increase probability of
            winning

    -   Receive greater utility than others
-   DEA technique
    -   Measure the relative efficiency of decision-making units that
        change inputs into outputs

    -   Compares the levels of inputs and outputs of one DMU with the
        rest of its peer group

        -   DMUs that produce the highest outputs with their inputs are
            deemed efficient

        -   Form a piecewise linear frontier

        -   Minimize inputs while satisfying at least the given output
            levels

            -   Searches data set to determine if a linear combination
                uses lower levels of inputs to produce at least the
                level of output
-   Use a two-input, three output BCC data envelopment analysis model
    -   MLB

        -   Inputs: hitter salaries and pitcher salaries

        -   Outputs: number of wins (includes postseason), team batting
            average, and team earned run average

        -   Looks at correlations between the variables

        -   A franchise is deemed inefficient only when other league
            franchise (or linear combination) can produce more wins,
            higher BA, lower ERA using lower pitcher and hitter salaries

    -   NFL

        -   Inputs: offense salary and defense salary

        -   Outputs: number of wins (includes postseason), offensive
            yards per game, defensive yards per game
-   1985 through 2001
-   Considers AL and NL separately
-   2001 Yankees one of the few franchises that appeared in the WS and
    found to be inefficient
-   28 out of 32 WS teams were perfectly efficient
-   19 out of 20 Super Bowl teams were perfectly efficient
-   NFL franchises more efficient on average than MLB
    -   significant difference in mean of the efficiency factors (both
        pre- and post-salary cap)
-   Significant increase in average efficiency after implementation of
    salary cap
-   Large-market teams are more likely to be inefficient and large
    market teams also more likely to have higher payrolls
-   No significant correlation between teams with high payrolls and
    inefficiency
-   No correlations between market size, payroll, and efficiency in the
    NFL
-   Uses glut ratio to compare payrolls and efficiency
    -   Glut ratio: franchise’s mean payroll ratio divided by its mean
        efficiency factor

    -   Inefficient franchise with a large payroll will have a high glut
        ratio

    -   Efficient franchise with a small payroll will have a low glut
        ratio
-   Strong significant correlation exists between large-market
    franchises and the glut ratio in MLB
    -   Large-market tended to be inefficient and have high payrolls

    -   Small-markets franchises tended to be efficient and have low
        payrolls
-   No statistically significant diff in avg efficiency between any year
    or set of years in MLB from 1985 to 2001
-   MLB economic infrastructure provides a significant advantage to
    big-market teams and encourages inefficient production
-   NFL has revenue-sharing system that compels owners to operate
    efficiently further encouraged by the implementation of a salary cap
-   Franchises with high payrolls in MLB have a greater chance of
    winning and are typically found to be efficient

## [Player Salaries, Organizational Efficiency, and Competitiveness in Major League Baseball](https://journals.sagepub.com/doi/epdf/10.1177/1527002506286776)

-   Research Questions
    -   How much does a team need to spend on TPS to be competitive?

        -   \$6.19 million in 1985 to \$38.67 million in 2002

        -   Annual growth rate of 10.7% adjusted for inflation

        -   Team minimum, mean, and maximum have risen at nearly the
            same rate

        -   Ratio of competitive salary to min TPS remained stable
            around mean of 1.5

    -   What is the maximum TPS that a team can pay without
        overspending?

        -   HC salary \$9.81 million in 1985 to \$125.9 million in 2002

        -   Increase of 12.6%

    -   How many teams are noncompetitive because of low TPS?

        -   Each year between 0 and 4

        -   Between 0 and 8 noncompetitive for other reasons

        -   Markov analysis suggests for a year, 5.5% of teams
            noncompetitive because of low TPS and another 4.8% of team
            noncompetitive for other reasons

    -   How many teams are overspending on TPS?

        -   Between 0 and 9 teams are HC

        -   95 of 442 (21.5%) have been overspending on TPS

        -   54.7% of overspending were HC, 43.2% were CC, and 2.15 were
            noncompetitive

        -   Markov analysis suggests for a year 12.2% of team will be HC

    -   How does noncompetitiveness because of low TPS relate to market
        size?

        -   Number of times a team is noncompetitive due to low TPS is
            negatively related to the size of the market

        -   Size of market relates to the number of games it can win if
            it is efficient

            -   Ex: New York team can expect to win about 4 more games
                per season than an efficient Milwaukee team

    -   How does overspending on TPS relate to market size?

        -   Large market teams are more likely to be HC than small
            market teams
-   DEA measures relative efficiency in situations with multiple inputs
    and outputs
    -   Establishes an efficient frontier based on the observed best
        performance
-   Logistic Regression: probability the dependent variable equals 1
    given specific values for input variable
-   Gini Index: scale from 0 to 1 to measure the extent to which members
    of a population are dissimilar from one another
-   Competitive balance: state where “every well-run club has regularly
    recurring reasonable hope of reaching postseason play”
-   Use a two-stage production model
    -   Stage 1: total player salary into talent measured by offense
        (total bases gained) and defensive (total bases surrendered)
        production

    -   Stage 2: talent into games won
-   Steps
    -   Solve DEA model for the front office of Team k

    -   Solve DEA model for the on-field performance of Team k

    -   Compute the hypothetical TBG and TBS that the team would have
        had had its front office been efficient

    -   Compute the number of games each team would have had they been
        efficient (EGW = GW/Efficiency)
-   Reasonable hope of reaching postseason play: team must have a least
    the same probability of reaching postseason play as it would have if
    all teams in its league were equally talented
    -   Refer to as team’s balanced probability

    -   Depends on the number of teams in its division and number of
        teams in its league
-   Construct a logistic regression for probability of qualifying for
    postseason play
    -   Computes probability of qualifying for postseason play given its
        number GW and the playoff qualification condition that applied
        to their league and division
-   Use logistic regression model to compute the minimum number of games
    a team must win to be competitive under each condition
    -   A team is competitive if and only if it would have won at least
        MGW had it been efficient
-   Determine minimum total player salary needed to be competitive
    -   Sort teams according to TPS (low to high) and use Gini index to
        identify a TPS that partitions teams into two sets

    -   Competitive salary for the year is the TPS for the lowest paid
        team in the primarily competitive set
-   Noncompetitive teams divided into two groups
    -   Low TPS or other reasons
-   Competitive teams into 3 groups
    -   Conditionally competitive if GW\* \< MGW

        -   Spending enough but inefficient in front office

        -   May be overspending if TPS\* \< TPS

    -   Economically competitive if GW\* \>= MGW and TPS\* = TPS

    -   Hypercompetitive if GW\* \>= MGW and TPS\* \< TPS
-   Model transition of teams among the 5 states according to a Markov
    process

# Next Steps

-   Use DEA analysis to look into market efficiency from 2015 through
    2024.
-   Inputs: Team Payroll
    -   Split by hitting vs pitching
        -   Potentially split further by position
        
    - Split by pre-arbitration, arbitration eligible, and veteran
-   Outputs: Wins, offensive statistic, and defensive statistic
    -   The first paper uses team batting average and team earned run
        average
    -   The second paper used total bases gain and total based
        surrendered
        -   I want to use different statistic

# Data Cleaning
## Import Data Sets 
```{r}
# Spotrac payroll data  
payroll_data <- read.csv("payroll_data_from_2015.csv") %>% select(-X)
payroll_ranks <- read.csv("payroll_ranks_from_2015.csv") %>% select(-X)

# Baseball Reference 40 man roster data
forty_man <- read.csv("bb_ref_forty_man_roster_from_2015.csv") %>% select(-X)

# Baseball Reference standings data
standings <- read.csv("standings_from_2012.csv") %>% select(-X) %>%
  filter(Tm != "Average")

# File that contains each team, there bbref abbr, league, and division info
team_league_map <- read.csv("team_league_map.csv") 
```


## Match League Info for Each Dataset 
```{r}
# Join standings data with team_league_map 
standings <- inner_join(standings, team_league_map, by = join_by(Tm == team))

# Join forty man data with team_league_map
# Only include the newest names to match spotrac
forty_man <- inner_join(
  forty_man,
  team_league_map %>%
    filter(team != "Cleveland Indians" &
             team != "Los Angeles Angels of Anaheim"),
  by = join_by(team == team_id)
) %>%
  rename(team_id = team, team = team.y)

# modify the format of the team name to match with spotrac 
team_league_map <- team_league_map %>% mutate(
  team = str_to_lower(team) %>%
    gsub(" ", "-", .) %>%
    gsub("\\.", "", .) %>%
    gsub("-of-anaheim", "", .) %>%
    gsub("indians", "guardians", .) %>%
    gsub("oakland-", "", .)
)  %>% distinct()

# Join payroll rank data on team_league_map
payroll_ranks <- inner_join(payroll_ranks, team_league_map, by = join_by(team))

# Join payroll data on team_league_map
payroll_data <- inner_join(payroll_data, team_league_map, by = join_by(team))

```

## General Data Cleaning
```{r}
# Clean payroll rank so amount and rank are numeric 
payroll_ranks <- payroll_ranks %>% mutate(
    amount = as.numeric(gsub("[\\$,]", "", amount)), 
    rank = as.numeric(gsub("[a-zA-Z]", "", rank))
  )

# Clean payroll data so salaries are numeric 
# Replace any empty strings with NA
# Simplify the roster column to not include year or the word "Payroll"
# This should match the designations found in the payroll_ranks
payroll_data <- payroll_data %>%
  mutate(across(
    c(
      payroll_salary,
      payroll_salary_adjusted,
      base_salary,
      signing_bonus,
      incentives_likely,
      incentives_unlikely
    ),
    ~ as.numeric(gsub("[\\$,]", "", na_if(., "-")))
  )) %>% 
  mutate(across(where(is.character), ~ na_if(.x, ""))) %>%
  mutate(
    roster = str_remove(roster, "^\\d{4} ") %>%  
                          str_remove(" Payroll$")
  )

# Determine total payroll for each team 
total_payroll <- payroll_ranks %>% 
  filter(player_type == "Total Adjusted Payroll Allocations")

# Clean 40 man data
# rename X.1 column to all_star
# Make Yrs numeric 
forty_man <- forty_man %>%
  rename(
    all_star = X.1
  ) %>%
  mutate(
    all_star = case_when(
      all_star == "All-Star" ~ TRUE, 
      TRUE ~ FALSE
    ), 
    Yrs = as.numeric(gsub("[a-zA-Z]", "", Yrs))
  )
  

# Add attribute for win variablity 
standings <- standings  %>%
  arrange(team_id, year) %>%  # Sort by team and year
  group_by(team_id) %>%       # Group by team
  mutate(win_variability = map_dbl(year, ~ {
    valid_years <- year[year <= .x & year != 2020]  # Exclude 2020
    if (length(valid_years) < 3) return(NA)  # Ensure at least 3 years exist
    sd(W[year %in% tail(valid_years, 3)])  # Compute SD for last 3 valid years
  })) %>%
  ungroup()
```

## Clean position attribute 
```{r}
# Add column to design hitter vs pitcher 
payroll_data <- payroll_data %>%
  mutate(
    pos_simple = case_when(
      pos == "SP/DH" ~ "two-way", 
      pos %in% c("RP", "SP", "SP1", "P", "CL") ~ "pitcher", 
      TRUE ~ "hitter"
    )
  )

# Assign a position to each player based on the most games played at a position
# For pitchers determine is starting or relief pitcher 
forty_man <- forty_man %>%
  mutate(
    position = apply(select(., P:DH) %>% select(-OF), 1, function(row)
      names(row)[which.max(row)]), 
    position = case_when(
      position == "P" & GS >= 0.8 * G ~ "SP", 
      position == "P" ~ "RP", 
      TRUE ~ position
    ), 
    position = gsub("X", "", position), 
    # manually fix Ohtani's position, don't love this solution 
    position = case_when(
      Name == "Shohei Ohtani" ~ "SP/DH", 
      TRUE ~ position
    ),
    pos_simple = case_when(
      position == "SP/DH" ~ "two-way", 
      position %in% c("RP", "SP") ~ "pitcher", 
      TRUE ~ "hitter"
    )
  )

```

## Handle Duplicate Players in Payroll 

The duplicates appear to come from when the player is on different roster designations throughout the season resulting in split contracts. To get the player's overall salary from the team I am going to add the duplicates together. For the other values, I want to keep it the same if they are the same in the duplicates, keep the non-null, and combine any remaining unmatched values. 

```{r}
# Function to combine duplicates into one row
combine_duplicates <- function(x) {
  if (all(is.na(x))) {
    return(NA)  # return NA if all values are NA
  } else if (is.numeric(x)) {
    return(sum(x, na.rm = TRUE))  # sum any numeric values
  } else if (is.character(x)) {
    return(paste(unique(na.omit(x)), collapse = "/"))  # concat diff values
  } else {
    return(unique(na.omit(x)))  # keep the value if the same between rows 
  }
}

payroll_data <- payroll_data %>% 
  group_by(player, team, year, pos_simple) %>%
  summarize(across(everything(), combine_duplicates), .groups = "drop")
```

## Match 40 man and payroll data

```{r}
# Function to clean name 
clean_name <- function(Name) {
  # Convert encoding to Latin-ASCII
  Name <- stri_trans_general(Name, "Latin-ASCII")
  
  # Remove suffixes
  suffixes <- c("Jr.", "Sr.", "II", "III", "IV", "V", "HOF")
  suffix_pattern <- paste0("\\s*(", paste(suffixes, collapse = "|"), ")\\s*$")
  Name <- stri_replace_all_regex(Name, suffix_pattern, "")
  #Name <- stri_replace_all_regex(Name, pattern, "")

  
  # Remove periods and dashes
  Name <- stri_replace_all_regex(Name, "[.]", "")
  
  Name <- stri_replace_all_regex(Name, "[-]", " ")
  
  # Convert to title case 
  Name <- str_to_title(Name)
  
  Name <- gsub('[\t\n]','',Name)
  
  return(Name)
}

# Function to extract last name while removing suffixes
extract_last_name <- function(full_name) {
  name_parts <- str_split(full_name, " ")[[1]]
  last_name <- tail(name_parts, 1)
  
  return(last_name)
}

match_players <- function(roster, payroll) {
  # Step 1: Clean the names by removing accents and create column for last name
  roster <- roster %>% mutate(
    player_name_clean = sapply(Name, clean_name), 
    last_name = sapply(player_name_clean, extract_last_name)
    )
  
  payroll <- payroll %>% mutate(
    player_name_clean = sapply(player, clean_name), 
    last_name = sapply(player_name_clean, extract_last_name)
    )
  
  # Step 2: Initial exact match on the cleaned names
  exact_matches <- roster %>%
    inner_join(payroll, by = 
                 join_by(player_name_clean, team_id, year, pos_simple), 
               suffix = c("", ""))
  
  # Step 3: Find unmatched rows in roster and payroll
  unmatched_roster <- roster %>%
    anti_join(exact_matches, by = join_by(player_name_clean, team_id, year, pos_simple))
  
  unmatched_payroll <- payroll %>%
    anti_join(exact_matches, by = join_by(player_name_clean, team_id, year, pos_simple))
  
  # Step 4: Fuzzy matching for unmatched names in roster
  last_name_matches <- unmatched_roster %>%
    inner_join(
      unmatched_payroll,
      by =join_by(last_name, team_id, year, pos_simple),
      suffix = c("", "")
    ) 
  
  combined_matches <- exact_matches %>%
    bind_rows(last_name_matches)
  
  unmatched_roster_2 <- unmatched_roster %>%
    anti_join(combined_matches, by = join_by(Name, team_id, year, pos_simple))
  
  unmatched_payroll_2 <- unmatched_payroll %>%
    anti_join(combined_matches, by = join_by(player, team_id, year, pos_simple))
  
  wo_position_matches <- unmatched_roster_2 %>%
    inner_join(unmatched_payroll_2, by = 
                 join_by(player_name_clean, team_id, year), 
               suffix = c("_bbref", "_spotrac")) %>% 
    rename(team = team_spotrac) %>%
    select(-contains("_spotrac"), -team_bbref) %>%
    rename_with(~ str_replace(.x, "_bbref$", ""), ends_with("_bbref"))
  
  combined_matches_2 <- combined_matches %>%
    bind_rows(wo_position_matches)
  
  unmatched_roster_3 <- unmatched_roster_2 %>%
    anti_join(combined_matches_2, by = join_by(Name, team_id, year, pos_simple))
  
  unmatched_payroll_3 <- unmatched_payroll_2 %>%
    anti_join(combined_matches_2, by = join_by(player, team_id, year, pos_simple))
  
  injured_players <- unmatched_payroll_3 %>% filter(roster == "Injured List")

  # Step 6: Combine exact matches, fuzzy matches, and remaining players from payroll
  # The remaining players are those who did not make a major league appearance for the team
  final_result <- combined_matches_2 %>%
    bind_rows(unmatched_roster_3)
  
  return(final_result)
}

roster_payroll_data <- match_players(forty_man, payroll_data)
```

```{r}

injured_players <- anti_join(payroll_data, roster_payroll_data,
                             by=join_by(player, team_id, year)) %>%
  filter(roster == "Injured List")

roster_to_join <- roster_payroll_data %>% 
  mutate(
    prev_year = year - 1, 
    next_year = year + 1
    ) %>%
  select(player, pos_simple, year, prev_year, next_year, Name, Age, Birth, B, T, Ht, Wt, DoB, Yrs) %>% distinct()

match_injured <- injured_players %>% 
  left_join(roster_to_join, 
            by = join_by(player, pos_simple, year  == next_year), 
            suffix = c("", "_matched")) %>%
  mutate(
    Age = Age + 1
  ) 

still_missing <- match_injured %>% filter(is.na(Name))

match_future <- still_missing %>% 
  left_join(roster_to_join, 
            by = join_by(player, pos_simple, year  == prev_year), 
            suffix = c("_original", "_matched")) %>%
  # mutate(
  #   Age = Age - 1, 
  #   Yrs = Yrs - 1
  # ) #%>%
  select(-contains("_original"))

final_missing <- match_future %>% filter(is.na(Name_matched))
  
```
# Initial Plots 

## Payroll Rank vs Team Performance 
```{r}
# Combine payroll data and standings
payroll_with_standings <- inner_join(
  total_payroll, standings, 
  by = join_by(team_id, year, league, division)
) 

# Payroll Rank vs Regular Season Rank
payroll_rank_reg_season_rank <- ggplot(payroll_with_standings, aes(x=rank, y=Rk)) + 
  geom_point() + 
  xlab("Payroll Rank") + ylab("Regular Season Rank") + 
  xlim(1, 30) + ylim(1,30) + 
  theme_bw() + ggtitle("Payroll vs Team Performance (Regular Season Rank)") + 
  theme(plot.title = element_text(hjust = 0.5))

ggsave("figures/payroll_rank_reg_season_rank.png", payroll_rank_reg_season_rank,
       width = 6, height = 4, units = "in")

payroll_rank_reg_season_rank

```

### By Year 

```{r}
# Payroll Rank vs Regular Season Rank split by Season 
payroll_rank_reg_season_rank_by_season <- ggplot(payroll_with_standings, aes(x=rank, y=Rk)) + 
  geom_point() + 
  xlab("Payroll Rank") + ylab("Regular Season Rank") + 
  xlim(1, 30) + ylim(1,30) + 
  theme_bw() + facet_wrap(vars(year), ncol=5) + 
  ggtitle("Payroll vs Team Performance (Regular Season Rank) by Season") + 
  theme(plot.title = element_text(hjust = 0.5))

ggsave("figures/payroll_rank_reg_season_rank_by_season.png", 
       payroll_rank_reg_season_rank_by_season,
       width = 10, height = 5, units = "in")

payroll_rank_reg_season_rank_by_season
```

## Payroll Amount vs Team Performance 
```{r}
# Payroll Amount vs Win Percentage
payroll_amount_win_per <- ggplot(payroll_with_standings, aes(x=amount, y=W)) + 
  geom_point() + 
  xlab("Payroll") + ylab("Wins") + 
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_bw() + ggtitle("Payroll vs Team Performance (Wins)") + 
  theme(plot.title = element_text(hjust = 0.5))

ggsave("figures/payroll_amount_win_per.png", 
       payroll_amount_win_per,
       width = 6, height = 4, units = "in")

payroll_amount_win_per

# Without 2020 
payroll_amount_win_per_wo_2020 <- ggplot(payroll_with_standings %>% 
                                           filter(year != 2020), aes(x=amount, y=W)) + 
  geom_point() + 
  xlab("Payroll") + ylab("Wins") + 
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_bw() + ggtitle("Payroll vs Team Performance (Wins) (Excluding 2020)") + 
  theme(plot.title = element_text(hjust = 0.5))

ggsave("figures/payroll_amount_win_per_wo_2020.png", 
       payroll_amount_win_per_wo_2020,
       width = 6, height = 4, units = "in")

payroll_amount_win_per_wo_2020
```

### By Year 
```{r}
# Payroll Amount vs Wins split by Season 
payroll_amount_wins_by_season <- ggplot(payroll_with_standings %>% 
                                          filter(year !=2020), aes(x=amount, y=W)) + 
  geom_point() + 
  xlab("Payroll") + ylab("Wins") + 
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) + 
  theme_bw() + facet_wrap(vars(year), ncol=5) + 
  ggtitle("Payroll vs Team Performance (Wins) by Season") + 
  theme(plot.title = element_text(hjust = 0.5))

ggsave("figures/payroll_amount_wins_by_season.png", 
       payroll_amount_wins_by_season,
       width = 10, height = 5, units = "in")

payroll_amount_wins_by_season


```

## Payroll Gap 
```{r}
payroll_gap <- payroll_with_standings %>%
  group_by(league, year) %>%
  summarize(
    payroll_gap = max(amount) - min(amount),
    win_gap = max(W) - min(W),
    .groups = "drop"
  ) %>%
  bind_rows(
    payroll_with_standings %>%
      group_by(year) %>%
      summarize(
        league = "Overall",
        payroll_gap = max(amount) - min(amount),
        win_gap = max(W) - min(W),
        .groups = "drop"
      )
  ) %>%
  mutate(
    gap_group = ifelse(year < 2020, "before_2020", "after_2020")
  )

payroll_gap_plot <- ggplot(payroll_gap, aes(x = year, group = league)) +
  geom_point(aes(y = payroll_gap, color = league)) +
  geom_line(aes(y = payroll_gap, color = league), linetype = 2) +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Season") +
  ylab("Payroll Gap") +
  labs(color = "League") +
  ggtitle("League Payroll Gap by Season") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x=element_text(angle=45, hjust=1))

win_gap_plot <- ggplot(payroll_gap, aes(x = year, group = league)) +
  geom_point(aes(y = win_gap, color = league)) +
  geom_line(aes(y = win_gap, color = league), linetype = 2) +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Season") +
  ylab("Win Gap") +
  labs(color = "League") +
  ggtitle("League Win Gap by Season") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x=element_text(angle=45, hjust=1))

payroll_win_gap <- ggarrange(payroll_gap_plot, win_gap_plot,
                             common.legend = TRUE, legend="bottom")

ggsave("figures/payroll_win_gap.png", 
       payroll_win_gap,
       width = 8, height = 4, units = "in")

payroll_win_gap


payroll_gap_plot_wo_2020 <- ggplot(payroll_gap %>% filter(year != 2020), 
                                   aes(x = year, 
                                       group = interaction(league, gap_group))) +
  geom_point(aes(y = payroll_gap, color = league)) +
  geom_line(aes(y = payroll_gap, color = league), linetype = 2) +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(breaks = 2015:2024) + 
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Season") +
  ylab("Payroll Gap") +
  labs(color = "League") +
  ggtitle("League Payroll Gap by Season") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x=element_text(angle=45, hjust=1))

win_gap_plot_wo_2020 <- ggplot(payroll_gap %>% filter(year != 2020), 
                               aes(x = year, 
                                   group = interaction(league, gap_group))) +
  geom_point(aes(y = win_gap, color = league)) +
  geom_line(aes(y = win_gap, color = league), linetype = 2) +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(breaks = 2015:2024) +
  theme_bw() +
  xlab("Season") +
  ylab("Win Gap") +
  labs(color = "League") +
  ggtitle("League Win Gap by Season") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x=element_text(angle=45, hjust=1))

payroll_win_gap_wo_2020 <- ggarrange(payroll_gap_plot_wo_2020, win_gap_plot_wo_2020, 
          common.legend = TRUE, legend="bottom")

ggsave("figures/payroll_win_gap_wo_2020.png", 
       payroll_win_gap_wo_2020,
       width = 8, height = 4, units = "in")

payroll_win_gap_wo_2020

```

## Min, Avg, Max Payroll 
```{r}
team_payroll_by_year <- total_payroll %>% group_by(league, year) %>%
  summarize(
    avg_team_payroll = mean(amount), 
    max_team_payroll = max(amount), 
    min_team_payroll = min(amount), 
    .groups = "drop"
  ) %>% bind_rows(
    total_payroll %>% 
      group_by(year) %>%
      summarize(
        league = "Overall", 
        avg_team_payroll = mean(amount), 
        max_team_payroll = max(amount), 
        min_team_payroll = min(amount), 
        .groups = "drop"
        )
  ) %>% pivot_longer(
    cols = c(avg_team_payroll, max_team_payroll, min_team_payroll), 
    names_to = "payroll_type",  
    values_to = "payroll_value" 
  ) %>% mutate(
    gap_group = ifelse(year < 2020, "before_2020", "after_2020")
  )

league_payroll_by_season <- ggplot(team_payroll_by_year, aes(x = year, 
                                 group = interaction(league, payroll_type))) +
  geom_point(aes(y = payroll_value, color = payroll_type)) +
  geom_line(aes(y = payroll_value, color = payroll_type, linetype = league)) +
  scale_color_brewer(palette = "Set1") +  
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) + 
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Season") +
  ylab("Team Payroll") +
  labs(color = "", linetype = "League") +
  ggtitle("League Payroll by Season") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x=element_text(angle=45, hjust=1))

ggsave("figures/league_payroll_by_season.png", 
       league_payroll_by_season,
       width = 8, height = 4, units = "in")

league_payroll_by_season

league_payroll_by_season_wo_2020 <- ggplot(team_payroll_by_year %>% filter(year != 2020), aes(x = year, 
                                 group = interaction(league, payroll_type, gap_group))) +
  geom_point(aes(y = payroll_value, color = payroll_type)) +
  geom_line(aes(y = payroll_value, color = payroll_type, linetype = league)) +
  scale_color_brewer(palette = "Set1") +  
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) + 
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Season") +
  ylab("Team Payroll") +
  labs(color = "", linetype = "League") +
  ggtitle("League Payroll by Season") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x=element_text(angle=45, hjust=1))

ggsave("figures/league_payroll_by_season_wo_2020.png", 
       league_payroll_by_season_wo_2020,
       width = 8, height = 4, units = "in")

league_payroll_by_season_wo_2020
```

### Split by Hitter vs Pitcher Payroll 

## Experience Level 
```{r}
roster_payroll_data %>% group_by(status) %>%
  summarise(
    count = n(), 
    avg_exp = mean(exp, na.rm = T), 
    na_exp = sum(is.na(exp)), 
    na_yrs = sum(is.na(Yrs))
  ) %>% arrange(desc(count))
```
```{r}
roster_payroll_data <- roster_payroll_data %>%
  mutate(
    status = case_when(
      (is.na(status) & !is.na(exp)) ~ ifelse(exp <= 3, "Pre-Arbitration", 
                                            ifelse(exp > 3 & exp <= 6, 
                                                   "Arbitration", "Veteran")), 
      (is.na(status) & is.na(exp) & 
         roster %in% 
         c("Minor", "Active Roster", "Injured List", "Retained/Minor")) ~ "Pre-Arbitration", 
       TRUE ~ status 
    )
  )
```

```{r}
roster_payroll_data %>% group_by(status) %>%
  summarise(
    count = n(), 
    avg_exp = mean(exp, na.rm = T), 
    na_exp = sum(is.na(exp)), 
    na_yrs = sum(is.na(Yrs))
  ) %>% arrange(desc(count))

```

```{r}
copy <- roster_payroll_data %>% select(player, DoB, year, exp, status) %>% 
  filter(!is.na(status)) %>% distinct()

status_lookup <- roster_payroll_data %>% filter(is.na(status) & is.na(exp)) %>% 
  left_join(copy, by = join_by(player, DoB, year), suffix = c("", "_matched")) %>%
    mutate(exp = ifelse(is.na(exp), exp_matched, exp), 
         status = ifelse(is.na(status), status_matched, status)
         ) %>%
  select(-contains("_matched")) %>% distinct()

roster_payroll_data <- bind_rows(roster_payroll_data %>% filter(!(is.na(status) & is.na(exp))), status_lookup) 
```


```{r}
copy <- roster_payroll_data %>% select(player, DoB, year, exp, status) %>% 
  filter(!is.na(status)) %>% distinct()

to_join <- roster_payroll_data %>% filter(is.na(status) & roster == "Retained") %>% mutate(
  prev_year = year - 1
) 

by_prev <- to_join %>% 
  left_join(copy, by = join_by(player, DoB, prev_year == year), suffix = c("", "_matched")) %>%
    mutate(prev_exp = ifelse(is.na(exp), exp_matched, exp), 
         prev_status = ifelse(is.na(status), status_matched, status), 
         status = case_when(
           (prev_status == "Veteran" | prev_exp >= 5) ~ "Veteran", 
           (prev_status == "Pre-Arbitration" & 
              (is.na(prev_exp) | prev_exp < 2)) ~ "Pre-Arbitration", 
           (is.na(prev_status) & is.na(prev_exp)) ~ 
             ifelse(Yrs <= 3, "Pre-Arbitration", ifelse(Yrs >= 8, "Veteran", "Arbitration")), 
           TRUE ~ "Arbitration"
         )
         ) %>%
  select(-contains("_matched")) %>% distinct()

roster_payroll_data <- bind_rows(roster_payroll_data %>% filter(!(is.na(status) &
                                                                    roster == "Retained")),
                                 roster_payroll_data %>% filter(is.na(player)),
                                 by_prev)
```


### Plots

```{r}
roster_payroll_data <- roster_payroll_data %>% 
  mutate(
    status_simple = case_when(
      grepl("Pre-Arbitration", status) ~ "Pre-Arbitration",
      (grepl("Veteran", status) | status == "Club Exercised") ~ "Veteran", 
      (grepl("Arbitration", status) | status == "Arb Avoided") ~ "Arbitration", 
      status == "minor-buried" ~ ifelse(exp < 3, "Pre-Arbitration", "Arbitration"), 
      exp < 3 ~ "Pre-Arbitration", 
      exp >= 3 & exp < 6 ~ "Arbitration", 
      exp >= 6 ~ "Veteran", 
      TRUE ~ status
    )
  )

# roster_payroll_data %>% group_by(status_simple) %>%
#   summarise(
#     count = n(), 
#     na_exp = sum(is.na(exp))
#   ) %>% arrange(desc(count))


status_summary <- roster_payroll_data %>% 
  filter(!is.na(status_simple) & !is.na(payroll_salary_adjusted)) %>%
  group_by(team_id, year, status_simple) %>%
  summarize(
    num_players = n(),
    min_salary = min(payroll_salary_adjusted, na.rm = T),
    max_salary = max(payroll_salary_adjusted, na.rm = T),
    avg_salary = mean(payroll_salary_adjusted, na.rm = T),
    total_payroll = sum(payroll_salary_adjusted, na.rm = T), 
    .groups = "drop"
  ) %>% 
  left_join(total_payroll %>% select(amount, rank, team_id, year, league, division, market_size), 
            by = join_by(team_id, year)) %>%
  rename(
    total_payroll_amount = amount, 
    total_payroll_rank = rank
  ) %>%
  left_join(standings %>% select(W, W.L., win_variability, team_id, year), 
            by = join_by(team_id, year))

#top <- test %>% group_by(status_simple) %>% slice_max(count, n = 10)


```

```{r}
status_summary <- status_summary %>% 
  mutate(status_simple = factor(status_simple, 
                                levels = c("Pre-Arbitration", "Arbitration", "Veteran")))

player_designation_distribution <- ggplot(status_summary %>% filter(year != 2020), aes(x = W, 
                                 group = interaction(status_simple))) +
  geom_point(aes(y = num_players, color = status_simple)) +
  theme_bw() +
  xlab("Wins") +
  ylab("Number of Players") +
  labs(color = "") +
  ggtitle("Player Designation Distribution") +
  theme(plot.title = element_text(hjust = 0.5))

ggsave("figures/player_designation_distribution.png", 
       player_designation_distribution,
       width = 8, height = 4, units = "in")

player_designation_distribution

payroll_player_distribution <- ggplot(status_summary %>% filter(year != 2020), aes(x = total_payroll_amount, 
                                 group = interaction(status_simple))) +
  geom_point(aes(y = avg_salary, color = status_simple)) +
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  scale_y_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Total Payroll Amount") +
  ylab("Avgerage Player Salary") +
  labs(color = "", linetype = "League") +
  ggtitle("Payroll Player Distribution") +
  theme(plot.title = element_text(hjust = 0.5))

ggsave("figures/payroll_player_distribution.png", 
       payroll_player_distribution,
       width = 8, height = 4, units = "in")

payroll_player_distribution

```

# DEA

```{r}
payroll_data %>% group_by(status) %>%
  summarise(
    count = n(), 
    avg_exp = mean(exp, na.rm = T), 
    na_exp = sum(is.na(exp))
  ) %>% arrange(desc(count))

missing_status <- payroll_data %>% filter(is.na(status))
```

```{r}
status_summary_wide <- status_summary %>% 
  pivot_wider(names_from = status_simple, 
              values_from = c(num_players, min_salary, max_salary, avg_salary, total_payroll), 
              names_glue = "{status_simple}_{.value}") %>%
  rename_with(~ str_to_lower(str_replace_all(., "-", "_"))) %>% 
  mutate(across(c("pre_arbitration_total_payroll", "arbitration_total_payroll", "veteran_total_payroll"), 
                 ~ . / 1e6, 
                .names = "{.col}_millions"))

```

```{r}
efficiency_data <- data.frame()

for (year in 2015:2024){
  dea_data <- make_deadata(status_summary_wide %>% filter(year == !!year), 
                           ni= 3, no = 1, dmus = "team_id", inputs = c("pre_arbitration_total_payroll_millions", "arbitration_total_payroll_millions", "veteran_total_payroll_millions"), outputs = "w")
  
  model <- model_basic(dea_data, rts = "vrs")

  efficiency <- data.frame(efficiencies(model))
  colnames(efficiency) <- c("efficiency")
  
  # Ensure team_id remains a character column
  efficiency <- efficiency %>%
    mutate(team_id = as.character(rownames(.)), year = year) %>%
    select(team_id, year, efficiency)  # Ensure correct column order
  
  rownames(efficiency) <- NULL
  
  efficiency_data <- bind_rows(efficiency_data, efficiency)
}


efficiency_data <- efficiency_data %>% left_join(status_summary_wide, by=join_by(team_id, year))

```


```{r}
dea_data <- make_deadata(status_summary_wide %>% filter(year == 2024), 
                         dmus = "team_id", 
                         inputs = c("pre_arbitration_total_payroll_millions", 
                                    "veteran_total_payroll_millions"), 
                         outputs = "w")
  
model <- model_basic(dea_data, rts = "vrs")

data.frame(efficiencies(model))

```

```{r}
ggplot(efficiency_data, aes(x = w.l.)) +
  geom_point(aes(y = efficiency)) +
  theme_bw() +
  xlab("Win Percentage") +
  ylab("Efficiency") +
  #labs(color = "", linetype = "League") +
  ggtitle("Wins vs Efficiency") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(efficiency_data , aes(x = total_payroll_amount)) +
  geom_point(aes(y = efficiency)) +
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Total Payroll") +
  ylab("Efficiency") +
  ggtitle("Payroll vs Efficiency") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(efficiency_data %>% filter(year != 2020) , aes(x = total_payroll_amount)) +
  geom_point(aes(y = efficiency)) +
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  theme_bw() +
  xlab("Total Payroll") +
  ylab("Efficiency") +
  ggtitle("Payroll vs Efficiency") +
  theme(plot.title = element_text(hjust = 0.5))

efficiency_data <- efficiency_data %>% mutate(eff_w = w / efficiency)
ggplot(efficiency_data , aes(x = w)) +
  geom_point(aes(y = eff_w)) +
  theme_bw() +
  xlab("Wins") +
  ylab("Efficient wins") +
  ggtitle("Games Won vs Efficient Games Won") +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r}
library(corrplot)
correlations <- cor(efficiency_data %>% select(-c(team_id, year, league, division)))

corrplot(correlations)
```

```{r}
correlations <- cor(efficiency_data %>% 
                      filter(year != 2020) %>% 
                      select(efficiency, total_payroll_amount, w, 
                             contains("_num_players"), contains("_total_payroll")))


corrplot(correlations, type = 'upper')
```

# Win Variance 
```{r}
status_summary_wide <- status_summary %>% 
  pivot_wider(names_from = status_simple, 
              values_from = c(num_players, min_salary, max_salary, avg_salary, total_payroll), 
              names_glue = "{status_simple}_{.value}") %>%
  rename_with(~ str_to_lower(str_replace_all(., "-", "_"))) %>%
  mutate(
    market_size_cat = case_when(
      market_size >= 5e6 ~ "large", 
      market_size < 5e6 & market_size >= 2.5e6 ~ "mid", 
      market_size < 2.5e6 ~ "small"
    ), 
    market_size_cat = factor(market_size_cat, levels = c("small", "mid", "large"))
  )
  

test <- status_summary_wide %>% 
  filter(year != 2020) %>% 
  group_by(team_id) %>%
  summarize(
    win_sd = sd(w), 
    total_payroll = sum(total_payroll_amount), 
    avg_payroll = mean(total_payroll_amount), 
    total_veteran_payroll = sum(veteran_total_payroll), 
    avg_veteran_payroll = mean(veteran_total_payroll), 
    total_arbitration_payroll = sum(arbitration_total_payroll), 
    avg_arbitration_payroll = mean(arbitration_total_payroll), 
    total_pre_arbitration_payroll = sum(pre_arbitration_total_payroll), 
    avg_pre_arbitration_payroll = mean(pre_arbitration_total_payroll), 
    market_size_cat = first(market_size_cat)
  )
  

ggplot(test, aes(x = avg_payroll, y = win_sd, color=market_size_cat)) + 
  geom_point() + 
  geom_text(aes(label = team_id), vjust = -0.5, size = 3) + 
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  ylim(0, max(test$win_sd)) + 
  xlab("Average Total Payroll Each Season") + 
  ylab("Wins") + 
  theme_bw()

ggplot(test, aes(x = avg_veteran_payroll, y = win_sd, color=market_size_cat)) + 
  geom_point() + 
  geom_text(aes(label = team_id), vjust = -0.5, size = 3) + 
  scale_x_continuous(labels = label_dollar(scale = 1/1e9, suffix = "B")) +
  ylim(0, max(test$win_sd)) + 
  xlab("Average Total Veteran Payroll Each Season") + 
  ylab("Wins") + 
  theme_bw()

ggplot(test, aes(x = avg_arbitration_payroll, y = win_sd, color=market_size_cat)) + 
  geom_point() + 
  geom_text(aes(label = team_id), vjust = -0.5, size = 3) + 
  scale_x_continuous(labels = label_dollar(scale = 1/1e9, suffix = "B")) +
  ylim(0, max(test$win_sd)) + 
  xlab("Average Total Arbitration Payroll Each Season") + 
  ylab("Wins") + 
  theme_bw()


ggplot(test, aes(x = avg_pre_arbitration_payroll, y = win_sd, color=market_size_cat)) + 
  geom_point() + 
  geom_text(aes(label = team_id), vjust = -0.5, size = 3) + 
  scale_x_continuous(labels = label_dollar(scale = 1/1e6, suffix = "M")) +
  ylim(0, max(test$win_sd)) + 
  xlab("Average Total Pre-Arbitration Payroll Each Season") + 
  ylab("Wins") + 
  theme_bw()
```


```{r}
status_summary_wo_2020 <- status_summary_wide %>% 
  filter(year != 2020) %>%
  group_by(team_id) %>%
  mutate(
    win_sd = sd(w)
  )

model <- lm(w ~ total_payroll_amount + team_id * as.factor(year), 
   data = status_summary_wo_2020)


summary(model)
```
```{r}
library(plm)
library(lmtest)

win_fe_mod <- plm(w ~ veteran_total_payroll + arbitration_total_payroll + pre_arbitration_total_payroll, 
                    data = status_summary_wo_2020,
                    index = c("team_id", "year"), 
                    model = "within")

summary(win_fe_mod)

#coeftest(win_fe_mod, vcov. = vcovHC, type = "HC1")
```

```{r}
win_market_fe_mod <- plm(w ~ total_payroll_amount + market_size, 
                         data = status_summary_wo_2020, 
                         index = c("team_id", "year"), 
                         model="within")

summary(win_market_fe_mod)

```

```{r}
df <- pdata.frame(status_summary_wo_2020, index = c("team_id", "year"))

fe_model <- plm(w ~ pre_arbitration_total_payroll + arbitration_total_payroll + veteran_total_payroll, 
                data = df, 
                model = "within", 
                effect = "individual")
summary(fe_model)

# re_model <- plm(w ~ pre_arbitration_total_payroll + arbitration_total_payroll + veteran_total_payroll, 
#                 data = df, 
#                 model = "random")
# summary(re_model)
# 
# phtest(fe_model, re_model)


```

```{r}
fe_model <- plm(win_variability ~ pre_arbitration_total_payroll + arbitration_total_payroll + veteran_total_payroll, 
                data = df, 
                model = "within", 
                effect = "individual")
summary(fe_model)
```